### 2013-01-14 paper.tex
first commit

%% bare_jrnl.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% see http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE!
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[journal]{IEEEtran}

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{remark}{Remark}

% to add comments in different colors (added by Azadeh)
\usepackage[usenames]{color}
\usepackage{lscape}
\usepackage{soul}
\newcommand{\blue}[1]{{\color{blue} #1}}
\newcommand{\red}[1]{{\color{red} #1}}
\newcommand{\black}[1]{{\color{black} #1}}

\newcommand{\Az}[1]{{\color{blue}{#1}}}
%\newcommand{\AzCom}[1]{{\it \color{magenta} [#1]}}
\newcommand{\AzCom}[1]{}
%\newcommand{\AzDel}[1]{{\color{red} \st{#1}}}
%\newcommand{\AzDel}[1]{{\color{Gray}{\{#1\}}}}
\newcommand{\AzDel}[1]{}

\newdimen\snellbaselineskip
\newdimen\snellskip
\snellskip=1.5ex
\snellbaselineskip=\baselineskip
\def\srule{\omit\kern.5em\vrule\kern-.5em}
\newbox\bigstrutbox
\setbox\bigstrutbox=\hbox{\vrule height14.5pt depth9.5pt width0pt}
\def\bigstrut{\relax\ifmmode\copy\bigstrutbox\else\unhcopy\bigstrutbox\fi}
\def\middlehrule#1#2{\noalign{\kern-\snellbaselineskip\kern\snellskip}
&\multispan#1\strut\hrulefill
&\omit\hbox to.5em{\hrulefill}\vrule
height \snellskip\kern-.5em&\multispan#2\hrulefill\cr}

\makeatletter
\def\bordermatrix#1{\begingroup \m@th
  \@tempdima 8.75\p@
  \setbox\z@\vbox{%
    \def\cr{\crcr\noalign{\kern2\p@\global\let\cr\endline}}%
    \ialign{$##$\hfil\kern2\p@\kern\@tempdima&\thinspace\hfil$##$\hfil
      &&\quad\hfil$##$\hfil\crcr
      \omit\strut\hfil\crcr\noalign{\kern-\snellbaselineskip}%
      #1\crcr\omit\strut\cr}}%
  \setbox\tw@\vbox{\unvcopy\z@\global\setbox\@ne\lastbox}%
  \setbox\tw@\hbox{\unhbox\@ne\unskip\global\setbox\@ne\lastbox}%
  \setbox\tw@\hbox{$\kern\wd\@ne\kern-\@tempdima\left(\kern-\wd\@ne
    \global\setbox\@ne\vbox{\box\@ne\kern2\p@}%
    \vcenter{\kern-\ht\@ne\unvbox\z@\kern-\snellbaselineskip}\,\right)$}%
  \null\;\vbox{\kern\ht\@ne\box\tw@}\endgroup}
\makeatletter

\makeatletter
\def\bordermatrix#1{\begingroup \m@th
  \@tempdima 8.75\p@
  \setbox\z@\vbox{%
    \def\cr{\crcr\noalign{\kern2\p@\global\let\cr\endline}}%
    \ialign{$##$\hfil\kern2\p@\kern\@tempdima&\thinspace\hfil$##$\hfil
      &&\quad\hfil$##$\hfil\crcr
      \omit\strut\hfil\crcr\noalign{\kern-\snellbaselineskip}%
      #1\crcr\omit\strut\cr}}%
  \setbox\tw@\vbox{\unvcopy\z@\global\setbox\@ne\lastbox}%
  \setbox\tw@\hbox{\unhbox\@ne\unskip\global\setbox\@ne\lastbox}%
  \setbox\tw@\hbox{$\kern\wd\@ne\kern-\@tempdima\left(\kern-\wd\@ne
    \global\setbox\@ne\vbox{\box\@ne\kern2\p@}%
    \vcenter{\kern-\ht\@ne\unvbox\z@\kern-\snellbaselineskip}\,\right)$}%
  \null\;\vbox{\kern\ht\@ne\box\tw@}\endgroup}
\makeatletter

%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

\usepackage{amssymb}
\usepackage{amsfonts}

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
% This package may be useful when used in conjunction with IEEEtran.cls'
% captionsoff option. Some IEEE journals/societies require that submissions
% have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.3.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% For subfigure.sty:
% \let\MYorigsubfigure\subfigure
% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat/subfig command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a
% page by themselves.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

\usepackage{psfrag}

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here

\def\Abf{{\mathbf{A}}}
\def\Pbf{{\mathbf{P}}}
\newcommand{\pr}[1]{\Pr \left\{#1\right\}}
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Modelling a Decentralized Constraint Satisfaction Solver for Collision-Free Channel Access}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{First~Author, %~\IEEEmembership{Member,~IEEE,}
        Second~Author, %~\IEEEmembership{Fellow,~OSA,}
        Third~Author, \IEEEmembership{Member, IEEE,}
        Fourth~Author, %~\IEEEmembership{Fellow,~OSA,}
        and~Last~Author,~\IEEEmembership{Senior~Member,~IEEE}% <-this % stops a space
\thanks{The authors are with Universitat Pompeu Fabra.
Roc Boronat 138, 08018 Barcelona, Catalunya, Spain.
E-mail: jaume.barcelo@upf.edu
}
}

% note the % following the last \IEEEmembership and also \thanks -
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
%
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~6, No.~1, January~2007}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
%
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}
\boldmath
As the physical layer (PHY) data rates increase and wireless local area network (WLAN) frames become shorter, the medium access control (MAC) overheads account for a larger fraction of the total transmission time.
This problem is known as the MAC bottleneck because it limits the throughput available to regardless no matter how advanced PHY techniques are used.
Current standardization efforts circumvent the problem by using aggregation.
By aggregating multiple network packets in a single MAC layer frame, the MAC overhead per useful network layer bit is reduced.

CSMA/ECA is an evolution of the well known CSMA/CA protocol that reduces the MAC contention overhead and therefore helps in battling the MAC bottleneck problem.
The core idea is the use of a deterministic backoff after successful transmissions.
In this paper we provide a survey of all the related contributions in the last years and describe the algorithm that can reach collision-free operation even for a large number of contenders.
We present simulation results to quantify the potential performance gains.

\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway.

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Medium Access Control, decentralized constraint satisfaction solver, learning MAC protocol
\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
%
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
%
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
%
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
%
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.

% \IEEEPARstart{I}{n} this paper we discuss a family of protocols that can be used for radio resource assignment in wireless networks.
% The problem of assigning resources in a distributed fashion is defined as a Decentralized Constraint Satisfaction (DCS) problem \cite{duffy2011dcs}.
% We consider a particular solver for this family of problems that works iteratively, in several rounds, to find a solution to the problem.

% The solver performs a stochastic search until it converges to a solution.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)

\IEEEPARstart{T}{his} is the introduction.
Blah blaah blahlah bababablah blalalah.
Blah blaah blahlah bababablah blalalah.
Blah blaah blahlah bababablah blalalah.

%\begin{figure}
%  \centering
%  \includegraphics[width=2.5in]{figures/csma_eca_compact}
%  \caption{CSMA/ECA contention}
%  \label{fig:csma_eca_compact}
%\end{figure}

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

%\appendices
%\section{Generalized Inclusion Exclusion Illustration}
%\label{app:incl-excl-thm}
%\label{app:theorem}
%We offer an example of the counting problem in (\ref{eq:Pij})  in which we are interested in computing the probability that exactly $delta$ among the $N$ stations succeed.
%
%\begin{figure}
%\psfrag{A1}[cc][cc]{$A_1$}
%\psfrag{A2}[cc][cc]{$A_2$}
%\psfrag{A3}[cc][cc]{$A_3$}
%\centering
%\includegraphics[height=3cm]{figures/counting}
%\caption{Illustration of the counting problem}
%\label{fig:counting}
%\end{figure}
%
%To illustrate the underlying idea we will use a simple example involving $N=3$ different stations choosing among $B=4$ different slots.
%We assume that the number of stations deterministically choosing its transmission slot is $d=0$ and we are interested in computing the probability that exactly $\delta=1$ station succeeds.
%
%There are a total of $B^N=4^3$ outcomes that are represented as black dots in Fig.~\ref{fig:counting}.
%Each outcome is equally likely with probability $1/64$.
%The figure also shows three events $A_1$, $A_2$ and $A_3$.
%The first event, $A_1$ represents the outcomes in which the first station succeeds.
%Similarly, the other two events, $A_2$ and $A_3$, represent the outcomes in which station 2 and 3 succeed, respectively.
%These sets are partially overlapping.
%Note also that it is impossible that two stations succeed while the third station collides.
%
%If we want to compute the probability that exactly one station succeeds, we have to count the outcomes that belong to either $A_1$, $A_2$ or $A_3$ and are not part of the intersections.
%This is $|A_1|+|A_2|+|A_3|-2(|A_1\cap A_2|+|A_1\cap A_3|+|A_2\cap A_3|)+3(|A_1\cap A_2\cap A_3|)$.
%In our paper we will use the intermediate values
% $S(1)=P(A_1)+P(A_2)+P(A_3)$,
% $S(2)=P(A_1\cap A_2)+P(A_1\cap A_3)+P(A_2\cap A_3)$, and $S(3)=P(A_1\cap A_2\cap A_3)$.
% The probability that exactly one station succeeds is finally computed  as $S(1)-2S(2)+3S(3)$.

% use section* for acknowledgement
%\section*{Acknowledgment}

%The authors would like to thank ...

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,my_bib}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}

%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.

%\end{thebibliography}

% biography section
%
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

%\begin{IEEEbiography}{Michael Shell}
%Biography text here.
%\end{IEEEbiography}
%
%% if you will not have a photo at all:
%\begin{IEEEbiographynophoto}{John Doe}
%Biography text here.
%\end{IEEEbiographynophoto}
%
%% insert where needed to balance the two columns on the last page with
%% biographies
%%\newpage
%
%\begin{IEEEbiographynophoto}{Jane Doe}
%Biography text here.
%\end{IEEEbiographynophoto}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2013-01-14 paper.tex
dunno

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Modelling a Decentralized Constraint Satisfaction Solver for Collision-Free Channel Access}
+\title{Battling the MAC Bottleneck in IEEE 802.11 Networks}
 %
@@
 As the physical layer (PHY) data rates increase and wireless local area network (WLAN) frames become shorter, the medium access control (MAC) overheads account for a larger fraction of the total transmission time.
-This problem is known as the MAC bottleneck because it limits the throughput available to regardless no matter how advanced PHY techniques are used.
+This problem is known as the MAC bottleneck because it limits the throughput available to the user no matter how advanced PHY techniques are used.
 Current standardization efforts circumvent the problem by using aggregation.

### 2013-01-22 paper.tex
before deleting everything

@@
-%% bare_jrnl.tex
+%% bare_conf.tex
 %% V1.3
@@
 %% by Michael Shell
-%% see http://www.michaelshell.org/
+%% See:
+%% http://www.michaelshell.org/
 %% for current contact information.
@@
 %% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
+%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
 %%
@@
 %% http://www.ieee.org/
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
 
@@
 
+% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
+% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
+% *** with production work. IEEE's font choices can trigger bugs that do  ***
+% *** not appear when using other class files.                            ***
+% The testflow support page is at:
+% http://www.michaelshell.org/tex/testflow/
+
 % Note that the a4paper option is mainly intended so that authors in
@@
 %
-\documentclass[journal]{IEEEtran}
-
-\newtheorem{definition}{Definition}
-\newtheorem{proposition}{Proposition}
-\newtheorem{theorem}{Theorem}
-\newtheorem{remark}{Remark}
-
-% to add comments in different colors (added by Azadeh)
-\usepackage[usenames]{color}
-\usepackage{lscape}
-\usepackage{soul}
-\newcommand{\blue}[1]{{\color{blue} #1}}
-\newcommand{\red}[1]{{\color{red} #1}}
-\newcommand{\black}[1]{{\color{black} #1}}
-
-\newcommand{\Az}[1]{{\color{blue}{#1}}}
-%\newcommand{\AzCom}[1]{{\it \color{magenta} [#1]}}
-\newcommand{\AzCom}[1]{}
-%\newcommand{\AzDel}[1]{{\color{red} \st{#1}}}
-%\newcommand{\AzDel}[1]{{\color{Gray}{\{#1\}}}}
-\newcommand{\AzDel}[1]{}
-
-\newdimen\snellbaselineskip
-\newdimen\snellskip
-\snellskip=1.5ex
-\snellbaselineskip=\baselineskip
-\def\srule{\omit\kern.5em\vrule\kern-.5em}
-\newbox\bigstrutbox
-\setbox\bigstrutbox=\hbox{\vrule height14.5pt depth9.5pt width0pt}
-\def\bigstrut{\relax\ifmmode\copy\bigstrutbox\else\unhcopy\bigstrutbox\fi}
-\def\middlehrule#1#2{\noalign{\kern-\snellbaselineskip\kern\snellskip}
-&\multispan#1\strut\hrulefill
-&\omit\hbox to.5em{\hrulefill}\vrule
-height \snellskip\kern-.5em&\multispan#2\hrulefill\cr}
-
-\makeatletter
-\def\bordermatrix#1{\begingroup \m@th
-  \@tempdima 8.75\p@
-  \setbox\z@\vbox{%
-    \def\cr{\crcr\noalign{\kern2\p@\global\let\cr\endline}}%
-    \ialign{$##$\hfil\kern2\p@\kern\@tempdima&\thinspace\hfil$##$\hfil
-      &&\quad\hfil$##$\hfil\crcr
-      \omit\strut\hfil\crcr\noalign{\kern-\snellbaselineskip}%
-      #1\crcr\omit\strut\cr}}%
-  \setbox\tw@\vbox{\unvcopy\z@\global\setbox\@ne\lastbox}%
-  \setbox\tw@\hbox{\unhbox\@ne\unskip\global\setbox\@ne\lastbox}%
-  \setbox\tw@\hbox{$\kern\wd\@ne\kern-\@tempdima\left(\kern-\wd\@ne
-    \global\setbox\@ne\vbox{\box\@ne\kern2\p@}%
-    \vcenter{\kern-\ht\@ne\unvbox\z@\kern-\snellbaselineskip}\,\right)$}%
-  \null\;\vbox{\kern\ht\@ne\box\tw@}\endgroup}
-\makeatletter
-
-\makeatletter
-\def\bordermatrix#1{\begingroup \m@th
-  \@tempdima 8.75\p@
-  \setbox\z@\vbox{%
-    \def\cr{\crcr\noalign{\kern2\p@\global\let\cr\endline}}%
-    \ialign{$##$\hfil\kern2\p@\kern\@tempdima&\thinspace\hfil$##$\hfil
-      &&\quad\hfil$##$\hfil\crcr
-      \omit\strut\hfil\crcr\noalign{\kern-\snellbaselineskip}%
-      #1\crcr\omit\strut\cr}}%
-  \setbox\tw@\vbox{\unvcopy\z@\global\setbox\@ne\lastbox}%
-  \setbox\tw@\hbox{\unhbox\@ne\unskip\global\setbox\@ne\lastbox}%
-  \setbox\tw@\hbox{$\kern\wd\@ne\kern-\@tempdima\left(\kern-\wd\@ne
-    \global\setbox\@ne\vbox{\box\@ne\kern2\p@}%
-    \vcenter{\kern-\ht\@ne\unvbox\z@\kern-\snellbaselineskip}\,\right)$}%
-  \null\;\vbox{\kern\ht\@ne\box\tw@}\endgroup}
-\makeatletter
-
+\documentclass[conference]{IEEEtran}
+% Add the compsoc option for Computer Society conferences.
 %
@@
 % manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
+% \documentclass[conference]{../sty/IEEEtran}
 
@@
 %
-\usepackage{cite}
+%\usepackage{cite}
 % cite.sty was written by Donald Arseneau
@@
   % driver is specified.
-  \usepackage[dvips]{graphicx}
+  % \usepackage[dvips]{graphicx}
   % declare the path(s) where your graphic files are
@@
   % every instance of \includegraphics
-  \DeclareGraphicsExtensions{.eps}
+  % \DeclareGraphicsExtensions{.eps}
 \fi
@@
 %
-\usepackage[cmex10]{amsmath}
+%\usepackage[cmex10]{amsmath}
 % A popular package from the American Mathematical Society that provides
@@
 % http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-\usepackage{amssymb}
-\usepackage{amsfonts}
 
@@
 
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
-% This package may be useful when used in conjunction with IEEEtran.cls'
-% captionsoff option. Some IEEE journals/societies require that submissions
-% have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.3.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% For subfigure.sty:
-% \let\MYorigsubfigure\subfigure
-% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat/subfig command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a
-% page by themselves.
-
 % *** PDF, URL AND HYPERLINK PACKAGES ***
 %
-\usepackage{url}
+%\usepackage{url}
 % url.sty was written by Donald Arseneau. It provides better support for
@@
 % \url{my_url_here}.
-
-\usepackage{psfrag}
 
@@
 % correct bad hyphenation here
-
-\def\Abf{{\mathbf{A}}}
-\def\Pbf{{\mathbf{P}}}
-\newcommand{\pr}[1]{\Pr \left\{#1\right\}}
 \hyphenation{op-tical net-works semi-conduc-tor}
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Battling the MAC Bottleneck in IEEE 802.11 Networks}
+\title{Battling the MAC Bottleneck in IEEE 802.11 WLANs}
+
+% author names and affiliations
+% use a multiple column layout for up to three different
+% affiliations
+\author{\IEEEauthorblockN{Michael Shell}
+\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
+Georgia Institute of Technology\\
+Atlanta, Georgia 30332--0250\\
+Email: http://www.michaelshell.org/contact.html}
+\and
+\IEEEauthorblockN{Homer Simpson}
+\IEEEauthorblockA{Twentieth Century Fox\\
+Springfield, USA\\
+Email: homer@thesimpsons.com}
+\and
+\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
+\IEEEauthorblockA{Starfleet Academy\\
+San Francisco, California 96678-2391\\
+Telephone: (800) 555--1212\\
+Fax: (888) 555--1212}}
+
+% conference papers do not typically use \thanks and this command
+% is locked out in conference mode. If really needed, such as for
+% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
+% after \documentclass
+
+% for over three affiliations, or if they all won't fit within the width
+% of the page, use this alternative format:
 % 
-%
-% author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
-
-\author{First~Author, %~\IEEEmembership{Member,~IEEE,}
-        Second~Author, %~\IEEEmembership{Fellow,~OSA,}
-        Third~Author, \IEEEmembership{Member, IEEE,}
-        Fourth~Author, %~\IEEEmembership{Fellow,~OSA,}
-        and~Last~Author,~\IEEEmembership{Senior~Member,~IEEE}% <-this % stops a space
-\thanks{The authors are with Universitat Pompeu Fabra.
-Roc Boronat 138, 08018 Barcelona, Catalunya, Spain.
-E-mail: jaume.barcelo@upf.edu
-}
-}
-
-% note the % following the last \IEEEmembership and also \thanks -
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-%
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
-
-% The paper headers
-\markboth{Journal of \LaTeX\ Class Files,~Vol.~6, No.~1, January~2007}%
-{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-%
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
+%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
+%Homer Simpson\IEEEauthorrefmark{2},
+%James Kirk\IEEEauthorrefmark{3}, 
+%Montgomery Scott\IEEEauthorrefmark{3} and
+%Eldon Tyrell\IEEEauthorrefmark{4}}
+%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
+%Georgia Institute of Technology,
+%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
+%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
+%Email: homer@thesimpsons.com}
+%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
+%Telephone: (800) 555--1212, Fax: (888) 555--1212}
+%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
 
@@
 \begin{abstract}
-\boldmath
+%\boldmath
 As the physical layer (PHY) data rates increase and wireless local area network (WLAN) frames become shorter, the medium access control (MAC) overheads account for a larger fraction of the total transmission time.
@@
 We present simulation results to quantify the potential performance gains.
-
 \end{abstract}
@@
 % This preserves the distinction between vectors and scalars. However,
-% if the journal you are submitting to favors bold math in the abstract,
+% if the conference you are submitting to favors bold math in the abstract,
 % then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals frown on math
-% in the abstract anyway.
+% of the abstract to achieve this. Many IEEE journals/conferences frown on
+% math in the abstract anyway.
 
-% Note that keywords are not normally used for peerreview papers.
-\begin{IEEEkeywords}
-Medium Access Control, decentralized constraint satisfaction solver, learning MAC protocol
-\end{IEEEkeywords}
+% no keywords
 
@@
 \section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-%
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-%
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-%
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-%
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
-
-% \IEEEPARstart{I}{n} this paper we discuss a family of protocols that can be used for radio resource assignment in wireless networks.
-% The problem of assigning resources in a distributed fashion is defined as a Decentralized Constraint Satisfaction (DCS) problem \cite{duffy2011dcs}.
-% We consider a particular solver for this family of problems that works iteratively, in several rounds, to find a solution to the problem.
-
-% The solver performs a stochastic search until it converges to a solution.
+% no \IEEEPARstart
+This demo file is intended to serve as a ``starter file''
+for IEEE conference papers produced under \LaTeX\ using
+IEEEtran.cls version 1.7 and later.
 % You must have at least 2 lines in the paragraph with the drop letter
 % (should never be an issue)
+I wish you the best of success.
 
-\IEEEPARstart{T}{his} is the introduction.
-Blah blaah blahlah bababablah blalalah.
-Blah blaah blahlah bababablah blalalah.
-Blah blaah blahlah bababablah blalalah.
+\hfill mds
 
-%\begin{figure}
-%  \centering
-%  \includegraphics[width=2.5in]{figures/csma_eca_compact}
-%  \caption{CSMA/ECA contention}
-%  \label{fig:csma_eca_compact}
+\hfill January 11, 2007
+
+\subsection{Subsection Heading Here}
+Subsection text here.
+
+\subsubsection{Subsubsection Heading Here}
+Subsubsection text here.
+
+% An example of a floating figure using the graphicx package.
+% Note that \label must occur AFTER (or within) \caption.
+% For figures, \caption should occur after the \includegraphics.
+% Note that IEEEtran v1.7 and later has special internal code that
+% is designed to preserve the operation of \label within \caption
+% even when the captionsoff option is in effect. However, because
+% of issues like this, it may be the safest practice to put all your
+% \label just after \caption rather than within \caption{}.
+%
+% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
+% option should be used if it is desired that the figures are to be
+% displayed while in draft mode.
+%
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=2.5in]{myfigure}
+% where an .eps filename suffix will be assumed under latex, 
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+%\caption{Simulation Results}
+%\label{fig_sim}
 %\end{figure}
 
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
+% Note that IEEE typically puts floats only at the top, even when this
+% results in a large percentage of a column being occupied by floats.
 
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
+% An example of a double column floating figure using two subfigures.
+% (The subfig.sty package must be loaded for this to work.)
+% The subfigure \label commands are set within each subfloat command, the
+% \label for the overall figure must come after \caption.
+% \hfil must be used as a separator to get equal spacing.
+% The subfigure.sty package works much the same way, except \subfigure is
+% used instead of \subfloat.
 %
+%\begin{figure*}[!t]
+%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
+%\label{fig_first_case}}
+%\hfil
+%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
+%\label{fig_second_case}}}
+%\caption{Simulation results}
+%\label{fig_sim}
+%\end{figure*}
+%
+% Note that often IEEE papers with subfigures do not employ subfigure
+% captions (using the optional argument to \subfloat), but instead will
+% reference/describe all of them (a), (b), etc., within the main caption.
 
-%\appendices
-%\section{Generalized Inclusion Exclusion Illustration}
-%\label{app:incl-excl-thm}
-%\label{app:theorem}
-%We offer an example of the counting problem in (\ref{eq:Pij})  in which we are interested in computing the probability that exactly $delta$ among the $N$ stations succeed.
+% An example of a floating table. Note that, for IEEE style tables, the 
+% \caption command should come BEFORE the table. Table text will default to
+% \footnotesize as IEEE normally uses this smaller font for tables.
+% The \label must come after \caption as always.
 %
-%\begin{figure}
-%\psfrag{A1}[cc][cc]{$A_1$}
-%\psfrag{A2}[cc][cc]{$A_2$}
-%\psfrag{A3}[cc][cc]{$A_3$}
+%\begin{table}[!t]
+%% increase table row spacing, adjust to taste
+%\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+%\caption{An Example of a Table}
+%\label{table_example}
 %\centering
-%\includegraphics[height=3cm]{figures/counting}
-%\caption{Illustration of the counting problem}
-%\label{fig:counting}
-%\end{figure}
-%
-%To illustrate the underlying idea we will use a simple example involving $N=3$ different stations choosing among $B=4$ different slots.
-%We assume that the number of stations deterministically choosing its transmission slot is $d=0$ and we are interested in computing the probability that exactly $\delta=1$ station succeeds.
-%
-%There are a total of $B^N=4^3$ outcomes that are represented as black dots in Fig.~\ref{fig:counting}.
-%Each outcome is equally likely with probability $1/64$.
-%The figure also shows three events $A_1$, $A_2$ and $A_3$.
-%The first event, $A_1$ represents the outcomes in which the first station succeeds.
-%Similarly, the other two events, $A_2$ and $A_3$, represent the outcomes in which station 2 and 3 succeed, respectively.
-%These sets are partially overlapping.
-%Note also that it is impossible that two stations succeed while the third station collides.
-%
-%If we want to compute the probability that exactly one station succeeds, we have to count the outcomes that belong to either $A_1$, $A_2$ or $A_3$ and are not part of the intersections.
-%This is $|A_1|+|A_2|+|A_3|-2(|A_1\cap A_2|+|A_1\cap A_3|+|A_2\cap A_3|)+3(|A_1\cap A_2\cap A_3|)$.
-%In our paper we will use the intermediate values
-% $S(1)=P(A_1)+P(A_2)+P(A_3)$,
-% $S(2)=P(A_1\cap A_2)+P(A_1\cap A_3)+P(A_2\cap A_3)$, and $S(3)=P(A_1\cap A_2\cap A_3)$.
-% The probability that exactly one station succeeds is finally computed  as $S(1)-2S(2)+3S(3)$.
+%% Some packages, such as MDW tools, offer better commands for making tables
+%% than the plain LaTeX2e tabular which is used here.
+%\begin{tabular}{|c||c|}
+%\hline
+%One & Two\\
+%\hline
+%Three & Four\\
+%\hline
+%\end{tabular}
+%\end{table}
+
+% Note that IEEE does not put floats in the very first column - or typically
+% anywhere on the first page for that matter. Also, in-text middle ("here")
+% positioning is not used. Most IEEE journals/conferences use top floats
+% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
+% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
+% command of the stfloats package.
+
+\section{Conclusion}
+The conclusion goes here.
+
+% conference papers do not normally have an appendix
 
 % use section* for acknowledgement
-%\section*{Acknowledgment}
+\section*{Acknowledgment}
 
-%The authors would like to thank ...
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
+The authors would like to thank...
 
@@
 % http://www.michaelshell.org/tex/ieeetran/bibtex/
-\bibliographystyle{IEEEtran}
+%\bibliographystyle{IEEEtran}
 % argument is your BibTeX string definitions and bibliography database(s)
-\bibliography{IEEEabrv,my_bib}
+%\bibliography{IEEEabrv,../bib/paper}
 %
@@
 % (used to reserve space for the reference number labels box)
-%\begin{thebibliography}{1}
+\begin{thebibliography}{1}
 
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
+\bibitem{IEEEhowto:kopka}
+H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
+  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
 
-%\end{thebibliography}
-
-% biography section
-%
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-%\begin{IEEEbiography}{Michael Shell}
-%Biography text here.
-%\end{IEEEbiography}
-%
-%% if you will not have a photo at all:
-%\begin{IEEEbiographynophoto}{John Doe}
-%Biography text here.
-%\end{IEEEbiographynophoto}
-%
-%% insert where needed to balance the two columns on the last page with
-%% biographies
-%%\newpage
-%
-%\begin{IEEEbiographynophoto}{Jane Doe}
-%Biography text here.
-%\end{IEEEbiographynophoto}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
+\end{thebibliography}
 

### 2013-01-22 paper.tex
new title, new paper

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Battling the MAC Bottleneck in IEEE 802.11 WLANs}
+\title{Evolution of CSMA Protocols fro the IEEE 802.11 Standard}
 
@@
 %\boldmath
-As the physical layer (PHY) data rates increase and wireless local area network (WLAN) frames become shorter, the medium access control (MAC) overheads account for a larger fraction of the total transmission time.
-This problem is known as the MAC bottleneck because it limits the throughput available to the user no matter how advanced PHY techniques are used.
-Current standardization efforts circumvent the problem by using aggregation.
-By aggregating multiple network packets in a single MAC layer frame, the MAC overhead per useful network layer bit is reduced.
-
-CSMA/ECA is an evolution of the well known CSMA/CA protocol that reduces the MAC contention overhead and therefore helps in battling the MAC bottleneck problem.
-The core idea is the use of a deterministic backoff after successful transmissions.
-In this paper we provide a survey of all the related contributions in the last years and describe the algorithm that can reach collision-free operation even for a large number of contenders.
-We present simulation results to quantify the potential performance gains.
+In this paper we present the requirements of candidate protocols to replace the pervasive CSMA/CA medium access control.
+We discuss the possibility of further preventing collisions and provide an overview of the related work.
+We specify protocols that are candidates of replacing CSMA/CA in pseudocode and use simulation to assess performance metrics such as throughput, fairness and collision probability.
 \end{abstract}
@@
 
-\hfill mds
+A candidate to replace CSMA/CA should 
+\begin{itemize}
+  \item Provide performance advantages, either in the form of throughput or short term fairness.
+  \item Be backward compatible with current implementation.
+  \item Be simple a simple evolution implementation to ease the transition and reduce time to market (Optional but desirable).
+\end{itemize}
+\section{Related Work}
 
-\hfill January 11, 2007
+\section{Enhanced CSMA}
 
-\subsection{Subsection Heading Here}
-Subsection text here.
-
-\subsubsection{Subsubsection Heading Here}
-Subsubsection text here.
+\section{Performance Evaluation}
 

### 2013-01-22 paper.tex
working on related work

@@
 %
-%\usepackage{cite}
+\usepackage{cite}
 % cite.sty was written by Donald Arseneau
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Evolution of CSMA Protocols fro the IEEE 802.11 Standard}
+\title{Evolution of CSMA Protocols of the IEEE 802.11 Standard}
 
@@
 \section{Related Work}
+Since the popularization of IEEE 802.11, several papers have proposed modifications to the contention protocol that is used for sharing the medium.
+They can be categorized in three groups regarding the approach they use.
+The first one prevents that the contention window is reset to its minimum value after successful transmissions.
+Examples of this first group include \cite{bharghavan1994map,wang2004ncr}.
+This solution improves throughput in saturation conditions at the price of lowering short term fairness.
+
+The second groups involves the accurate estimation of the number of contenders to adjust the contention parameters.
+Two examples of this group are \cite{cali2000dti,lopez-toledo2006aoi}.
+This approach offers some throughput and fairness gains at the expense of increased implementation complexity.
+As the number of contenders is estimated relying on the number of collisions, the presence of channel errors further complicates the estimation.
+Furthermore, there is a fundamental trade-off between the accuracy and reaction time of the estimation.
+
+The aforementioned solutions are not able to fairly share the medium with legacy devices.In fact, these proposals are, generally speaking, less aggressive than the currently implemented protocol.
+Consequently, in a shared network, the new stations would receive a smaller share of the available bandwidth in a mixed scenario.
+
+A more important limitation of the solutions exposed so far is that the throughput is bounded by that of CSMA/CA with optimal configuration \cite{cali2000dti,bianchi2000pai}.
+In this paper we focus on a third group of solutions that delivers throughput above the maximum attainable by CSMA/CA.
+
+This third group of solutions uses a deterministic backoff after successful transmissions to further reduce the chances of collisions.
+It was introduced in \cite{barcelo2008lba} and a more detailed analysis that includes both saturated and non-saturated conditions is presented in \cite{bellalta2009vtc}.
+analyzed in depth in \cite{he2009srb}.
 
@@
 % http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
+\bibliographystyle{IEEEtran}
 % argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
+\bibliography{IEEEabrv,my_bib}
 %
@@
 % (used to reserve space for the reference number labels box)
-\begin{thebibliography}{1}
-
-\bibitem{IEEEhowto:kopka}
-H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-\end{thebibliography}
+%\begin{thebibliography}{1}
+%
+%\bibitem{IEEEhowto:kopka}
+%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
+%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
+%
+%\end{thebibliography}
 

### 2013-01-22 paper.tex
finished related work (?)

@@
 This third group of solutions uses a deterministic backoff after successful transmissions to further reduce the chances of collisions.
+Under certain conditions, collision-free operation is reached.
 It was introduced in \cite{barcelo2008lba} and a more detailed analysis that includes both saturated and non-saturated conditions is presented in \cite{bellalta2009vtc}.
-analyzed in depth in \cite{he2009srb}.
+A more in-depth study is carried out in \cite{he2009srb}, including realistic elements such as clear channel assessment errors.
+Different aspects of fairness are addressed in \cite{he2009srb,barcelo2010fcc,fang2011dlm}.
+The performance in realistic channels taking into account the Auto Rate Fallback mechanism is evaluated \cite{martorell2012pec,martorell2012tfl}.
+
+Even though initial research efforts where focused on the WLAN collision problem, more recent papers try to extend the idea to multi-hop networks.
+In \cite{hui2011epp}, the multi-hop slotted case explored.
+The more realistic situation in multi-hop networks in which the time is not slotted is studied in \cite{barcelo2013dcc}.
+
+The same principles that we exploit to prevent collisions in WLANs can be used in other areas of radio resource management in wireless area networks \cite{duffy2011dcs,checco2012scs,checco2012lbc,khan2013aso}.
+
+In all the work cited so far, there is a limitation that the number of contenders should not exceed the value of the deterministic backoff used after successful transmissions.
+If this value is exceeded, it is no longer possible to achieve collision-free operation.
+A first solution to solve this problem is presented in \cite{barcelo2011tcf}, but it requires the presence of a central entity (typically the access point) that instructs the other nodes to adjust the value of their deterministic backoff.
+
+In the present paper, we study a completely distributed solution to accommodate a large number of contenders in a fair collision-free fashion.
 

### 2013-01-22 paper.tex
a round of typos

@@
 As the number of contenders is estimated relying on the number of collisions, the presence of channel errors further complicates the estimation.
-Furthermore, there is a fundamental trade-off between the accuracy and reaction time of the estimation.
+Furthermore, there is a fundamental trade-off between the accuracy and the reaction time of the estimation.
 
-The aforementioned solutions are not able to fairly share the medium with legacy devices.In fact, these proposals are, generally speaking, less aggressive than the currently implemented protocol.
-Consequently, in a shared network, the new stations would receive a smaller share of the available bandwidth in a mixed scenario.
+The aforementioned solutions are not able to fairly share the medium with legacy devices.
+In fact, these proposals are, generally speaking, less aggressive than the currently implemented protocol.
+Consequently, in a hypothetical mixed network in which the new and old protocols coexist, the new stations would receive a smaller share of the available bandwidth in a scenario.
 
 A more important limitation of the solutions exposed so far is that the throughput is bounded by that of CSMA/CA with optimal configuration \cite{cali2000dti,bianchi2000pai}.
-In this paper we focus on a third group of solutions that delivers throughput above the maximum attainable by CSMA/CA.
+In the present paper we focus on a third group of solutions that delivers throughput above the maximum attainable by CSMA/CA.
 
@@
 
-In all the work cited so far, there is a limitation that the number of contenders should not exceed the value of the deterministic backoff used after successful transmissions.
+In all the previous work on collision-free operation in WLANs mentioned so far, there is the limitation that the number of contenders should not exceed the value of the deterministic backoff used after successful transmissions.
 If this value is exceeded, it is no longer possible to achieve collision-free operation.

### 2013-01-23 paper.tex
working on algorithms
algorithms with end
working on the algorithms
added fair-share
ready to rewrite algorithms

@@
 
+%\usepackage[linesnumbered,noend]{algorithm2e}
+\usepackage[linesnumbered]{algorithm2e}
+
 % *** Do not adjust lengths that control margins, column widths, etc. ***
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Evolution of CSMA Protocols of the IEEE 802.11 Standard}
+\title{Evolution of CSMA Protocols for the IEEE 802.11 Standard}
 
@@
 
+\begin{algorithm}
+  %\tcc{Initialize $b$.}
+  $b \leftarrow \mathcal{U}[0,\rm{CW}_{min}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    $a \leftarrow 0$ \;
+    \While{$a < A$}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot \;
+        $b \leftarrow b-1$ \;
+      }
+      Attempt transmission \;
+      \eIf{success}{
+        \tcc{Random backoff.}
+        $b \leftarrow \mathcal{U}[0,\rm{CW}_{min}-1]$\;
+        break \;
+      }
+      {
+        \tcc{Random backoff.}
+        $a \leftarrow a+1$ \;
+        $b \leftarrow \mathcal{U}[0,\min (\rm{CW}_{min}*2^a,\rm{CW}_{max}) -1]$\;
+      }
+    }
+  }
+  \caption{CSMA/CA}
+  \label{alg:CSMA_ECA}
+\end{algorithm}
+
+\begin{algorithm}
+  %\tcc{Initialize $b$.}
+  $b \leftarrow \mathcal{U}[0,\rm{CW}_{min}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    $a \leftarrow 0$ \;
+    \While{$a < A$}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot \;
+        $b \leftarrow b-1$ \;
+      }
+      Attempt transmission \;
+      \eIf{success}{
+        \tcc{Deterministic backoff.}
+        $b \leftarrow \rm{CW}_{min}/2$\;
+        break \;
+      }
+      {
+        %\tcc{If transmission fails.}
+        $a \leftarrow a+1$ \;
+        \tcc{fall to random backoff.}
+        $b \leftarrow \mathcal{U}[0,\min (\rm{CW}_{min}*2^a,\rm{CW}_{max}) -1]$\;
+      }
+    }
+  }
+  \caption{CSMA/ECA}
+  \label{alg:CSMA_ECA}
+\end{algorithm}
+
+\begin{algorithm}
+  %\tcc{Initialize $b$.}
+  $b \leftarrow \mathcal{U}[0,\rm{CW}_{min}-1]$\;
+  \tcc{Hysteresis: The backoff stage is reset only when a node joins the contention or the queue is empty.}
+  $a \leftarrow 0$ \;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \While{$a < A$}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot \;
+        $b \leftarrow b-1$ \;
+      }
+      Attempt transmission \;
+      \eIf{success}{
+        \tcc{Deterministic backoff.}
+        $b \leftarrow \rm{CW}_{min}/2$\;
+        break \;
+      }
+      {
+        %\tcc{If transmission fails.}
+        $a \leftarrow a+1$ \;
+        \tcc{fall to random backoff.}
+        $b \leftarrow \mathcal{U}[0,\min (\rm{CW}_{min}*2^a,\rm{CW}_{max}) -1]$\;
+        \If{$a > m$}
+        {
+          Discard packet \;
+          break \;
+        }
+      }
+    }
+  }
+  \caption{CSMA/ECA with hysteresis}
+  \label{alg:CSMA_ECA_hist}
+\end{algorithm}
+
+\begin{algorithm}
+  %\tcc{Initialize $b$.}
+  $b \leftarrow \mathcal{U}[0,\rm{CW}_{min}-1]$\;
+  \tcc{Hysteresis: The backoff stage is reset only when a node joins the contention or the queue is empty.}
+  $a \leftarrow 0$ \;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \While{$a < A$}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot \;
+        $b \leftarrow b-1$ \;
+      }
+      \tcc{Fair-share: $2^a$ packets are transmitted.}
+      Attempt aggregate transmission of $2^a$ packets\;
+      \eIf{success}{
+        \tcc{Deterministic backoff.}
+        $b \leftarrow \rm{CW}_{min}/2$\;
+        break \;
+      }
+      {
+        %\tcc{If transmission fails.}
+        $a \leftarrow a+1$ \;
+        \tcc{fall to random backoff.}
+        $b \leftarrow \mathcal{U}[0,\min (\rm{CW}_{min}*2^a,\rm{CW}_{max}) -1]$\;
+      }
+    }
+  }
+  \caption{CSMA/ECA with hysteresis and fair-share}
+  \label{alg:CSMA_ECA_hist}
+\end{algorithm}
+
 \section{Performance Evaluation}

### 2013-01-23 paper.tex
I should consider the special case a=R
new csma/ca algorithm
adjusted to Boris suggestions for algorithm
new version of algorithms ready
brief description of the algorithms

--- 2013-01-22 paper.tex
@@
 
+%\usepackage[linesnumbered,noend]{algorithm2e}
+\usepackage[linesnumbered,vlined]{algorithm2e}
+
 % *** Do not adjust lengths that control margins, column widths, etc. ***
@@
 
+\begin{algorithm}
+\While{True}
+{
+  %\tcc{Initialize retransmission attempts.}
+  $r \leftarrow 0$ \;
+  %\tcc{Initialize backoff stage.}
+  $s \leftarrow 0$ \;
+  %\tcc{Initialize backoff counter.}
+  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot \;
+        $b \leftarrow b-1$ \;
+      }
+      Attempt transmission \;
+      \eIf{success}{
+        %\tcc{Random backoff.}
+        $r \leftarrow 0$ \;
+        $s \leftarrow 0$ \;
+        $b \leftarrow \mathcal{U}[0,2^{s}\rm{CW}_{min}-1]$\;
+      }
+      {
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$ \;
+        $s \leftarrow \min r(s+1,S)$ \;
+        $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min} -1]$\;
+      }
+    }
+    \If{r=R}
+    {
+      Discard packet\;
+      $r \leftarrow 0$ \;
+      $s \leftarrow 0$ \;
+      $b \leftarrow \mathcal{U}[0,2^s \rm{CW}_{min}-1]$\;
+    }
+  }
+}
+\caption{CSMA/CA}
+\label{alg:CSMA_CA}
+\end{algorithm}
+
+\begin{algorithm}
+\While{True}
+{
+  %\tcc{Initialize retransmission attempts.}
+  $r \leftarrow 0$ \;
+  %\tcc{Initialize backoff stage.}
+  $s \leftarrow 0$ \;
+  %\tcc{Initialize backoff counter.}
+  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot \;
+        $b \leftarrow b-1$ \;
+      }
+      Attempt transmission \;
+      \eIf{success}{
+        %\tcc{Random backoff.}
+        $r \leftarrow 0$ \;
+        $s \leftarrow 0$ \;
+        $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
+      }
+      {
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$ \;
+        $s \leftarrow \min r(s+1,S)$ \;
+        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
+      }
+    }
+    \If{r=R}
+    {
+      Discard packet\;
+      $r \leftarrow 0$ \;
+      $s \leftarrow 0$ \;
+      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
+    }
+  }
+}
+\caption{CSMA/ECA}
+\label{alg:CSMA_ECA}
+\end{algorithm}
+
+\begin{algorithm}
+\While{True}
+{
+  %\tcc{Initialize retransmission attempts.}
+  $r \leftarrow 0$ \;
+  %\tcc{Initialize backoff stage.}
+  $s \leftarrow 0$ \;
+  %\tcc{Initialize backoff counter.}
+  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot \;
+        $b \leftarrow b-1$ \;
+      }
+      Attempt transmission \;
+      \eIf{success}{
+        %\tcc{Random backoff.}
+        $r \leftarrow 0$ \;
+        %$s \leftarrow 0$ \;
+        $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
+      }
+      {
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$ \;
+        $s \leftarrow \min r(s+1,S)$ \;
+        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
+      }
+    }
+    \If{r=R}
+    {
+      Discard packet\;
+      $r \leftarrow 0$ \;
+      %$s \leftarrow 0$ \;
+      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
+    }
+  }
+}
+\caption{CSMA/ECA with hysteresis}
+\label{alg:CSMA_ECA_hist}
+\end{algorithm}
+
+\begin{algorithm}
+\While{True}
+{
+  %\tcc{Initialize retransmission attempts.}
+  $r \leftarrow 0$ \;
+  %\tcc{Initialize backoff stage.}
+  $s \leftarrow 0$ \;
+  %\tcc{Initialize backoff counter.}
+  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot \;
+        $b \leftarrow b-1$ \;
+      }
+      Attempt transmission of $2^s$ packets \;
+      \eIf{success}{
+        %\tcc{Random backoff.}
+        $r \leftarrow 0$ \;
+        %$s \leftarrow 0$ \;
+        $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
+      }
+      {
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$ \;
+        $s \leftarrow \min r(s+1,S)$ \;
+        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
+      }
+    }
+    \If{r=R}
+    {
+      Discard packet\;
+      $r \leftarrow 0$ \;
+      %$s \leftarrow 0$ \;
+      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
+    }
+  }
+}
+\caption{CSMA/ECA with hysteresis and fair-share}
+\label{alg:CSMA_ECA_hist_fair}
+\end{algorithm}
+
+Algorithm \ref{alg:CSMA_CA} describes the CSMA/CA algorithm that is used in current networks.
+When a station joins the contention it initilalizes the retry attempt counter $r$ and the backoff stage $s$ to zero.
+The backoff counter $b$ is initialized using a uniform random distribution and a contention window.
+After each collision, the retry attempt counter and the backoff stage counter are incremented.
+As a consecuence of the incremented backoff stage, a larger contention window is used.
+Note that there is maximum backoff stage $S$ and a maximum retry limit $R$.
+When the maximum retry  limit is reached the packet is discarded, $r$ and $s$ are reset anb a new value for $b$ is computed.
+Similarly, if the packet is successfully transmitted, the $s$ and $r$ counters are reset and a new random backoff counter is computed.
+
+Algorithm \ref{alg:CSMA_ECA} describes CSMA/ECA in which a deterministic backoff is used after successful transmission.
+The only change with respect to CSMA/CA is in line 14, where the determimistic assignment takes place.
+
+Hysteresis is included in \ref{alg:CSMA_ECA_hist}.
+Adding hysteresis is as simple as removing lines 13 and 23 from Algoritm \ref{alg:CSMA_ECA}.
+
+Finally, fair-share is implemented in \ref{alg:CSMA_ECA_hist_fair}.
+Line 10 of \ref{alg:CSMA_ECA_hist} is modified to increas the number of transmitted packets from one to $2^s$.
+
 \section{Performance Evaluation}

### 2013-01-24 paper.tex
authors

@@
 % affiliations
-\author{\IEEEauthorblockN{Michael Shell}
-\IEEEauthorblockA{School of Electrical and\\Computer Engineering\\
-Georgia Institute of Technology\\
-Atlanta, Georgia 30332--0250\\
-Email: http://www.michaelshell.org/contact.html}
-\and
-\IEEEauthorblockN{Homer Simpson}
-\IEEEauthorblockA{Twentieth Century Fox\\
-Springfield, USA\\
-Email: homer@thesimpsons.com}
-\and
-\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
-\IEEEauthorblockA{Starfleet Academy\\
-San Francisco, California 96678-2391\\
-Telephone: (800) 555--1212\\
-Fax: (888) 555--1212}}
+\author{\IEEEauthorblockN{Luis Sanabria-Russo, Azadeh Faridi, Boris Bellalta, Jaume Barcelo, Miquel Oliver}
+\IEEEauthorblockA{
+Universitat Pompeu Fabra\\
+Roc Boronat 183, 08018 Barcelona\\
+Email: \{name.surname@upf.edu\}}
+}
+%\and
+%\IEEEauthorblockN{Homer Simpson}
+%\IEEEauthorblockA{Twentieth Century Fox\\
+%Springfield, USA\\
+%Email: homer@thesimpsons.com}
+%\and
+%\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
+%\IEEEauthorblockA{Starfleet Academy\\
+%San Francisco, California 96678-2391\\
+%Telephone: (800) 555--1212\\
+%Fax: (888) 555--1212}}
 
@@
   \item Be backward compatible with current implementation.
+  \item Support a large number of simultaneous contenders.
   \item Be simple a simple evolution implementation to ease the transition and reduce time to market (Optional but desirable).
@@
 
+In this section we describe the CSMA/CA protocol as it is currently implemented in the IEEE 802.11 and an evolution of the standard that satisfies the four requirements specified in the introduction.
+Therefore, the presented protocol is a valid candidate to replace CSMA/CA in the upcoming revisions of the standard.
+
+There are three changes in the new protocol compared to the legacy one.
+Firstly, a deterministic backoff is used after successful transmissions.
+Secondly, the backoff stage is not reset after a packet is serviced.
+The backoff stage is reset only when the station leaves the contention because it has no packet to be transmitted.
+And thirdly, the number of packets transmitted in every transmission attempt is a function of the backoff stage.
+Note that current standards already support the transmission of multiple packets in a single slot.
+
+In the following we describe the original protocol and then we describe each of the modifications that we propose, one by one.
+
 Algorithm \ref{alg:CSMA_CA} describes the CSMA/CA algorithm that is used in current networks.
-When a station joins the contention it initilalizes the retry attempt counter $r$ and the backoff stage $s$ to zero.
+When a station joins the contention it initializes the retry attempt counter $r$ and the backoff stage $s$ to zero.
 The backoff counter $b$ is initialized using a uniform random distribution and a contention window.
 After each collision, the retry attempt counter and the backoff stage counter are incremented.
-As a consecuence of the incremented backoff stage, a larger contention window is used.
+As a consequence of the incremented backoff stage, a larger contention window is used.
 Note that there is maximum backoff stage $S$ and a maximum retry limit $R$.
@@
 Hysteresis is included in \ref{alg:CSMA_ECA_hist}.
-Adding hysteresis is as simple as removing lines 13 and 23 from Algoritm \ref{alg:CSMA_ECA}.
+Adding hysteresis is as simple as removing lines 13 and 23 from Algorithm \ref{alg:CSMA_ECA}.
+Note that when hysteresis is used, the backoff stage is reset only when the node has no packet to serve.
 
 Finally, fair-share is implemented in \ref{alg:CSMA_ECA_hist_fair}.
-Line 10 of \ref{alg:CSMA_ECA_hist} is modified to increas the number of transmitted packets from one to $2^s$.
+Line 10 of \ref{alg:CSMA_ECA_hist} is modified to increase the number of transmitted packets from one to $2^s$.
 

### 2013-01-24 paper.tex
enhanced csma/eca
removed r
changed abstract

@@
 We discuss the possibility of further preventing collisions and provide an overview of the related work.
-We specify protocols that are candidates of replacing CSMA/CA in pseudocode and use simulation to assess performance metrics such as throughput, fairness and collision probability.
+We specify a protocol that is a candidate for replacing CSMA/CA in pseudocode and use simulation to assess performance metrics such as throughput, fairness and collision probability.
 \end{abstract}
@@
 
-\section{Enhanced CSMA}
+\section{Enhanced CSMA/CA}
 
@@
         $r \leftarrow r+1$ \;
-        $s \leftarrow \min r(s+1,S)$ \;
+        $s \leftarrow \min (s+1,S)$ \;
         $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min} -1]$\;
@@
         $r \leftarrow r+1$ \;
-        $s \leftarrow \min r(s+1,S)$ \;
+        $s \leftarrow \min (s+1,S)$ \;
         $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;

### 2013-01-24 paper.tex
azadeh s suggestions

@@
 \begin{algorithm}
-\While{True}
+\While{the device is on}
 {
@@
       }
-      Attempt transmission \;
+      Attempt transmission of 1 packet \;
       \eIf{success}{
@@
   }
+  Wait for the next packet \;
 }

### 2013-01-24 paper.tex
re-organized algorithms

@@
       Attempt transmission of 1 packet \;
+      \If{collision}{
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$ \;
+        $s \leftarrow \min (s+1,S)$ \;
+        $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min} -1]$\;
+      }
+    }
     \eIf{success}{
@@
     {
-        %\tcc{Random backoff.}
-        $r \leftarrow r+1$ \;
-        $s \leftarrow \min (s+1,S)$ \;
-        $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min} -1]$\;
-      }
-    }
-    \If{r=R}
-    {
       Discard packet\;
@@
   }
-  Wait for the next packet \;
+  Wait until there is a packet to transmit \;
 }
@@
       Attempt transmission of 1 packet \;
+      \If{collision}
+      {
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$ \;
+        $s \leftarrow \min (s+1,S)$ \;
+        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
+      }
+    }
     \eIf{success}{
@@
     {
-        %\tcc{Random backoff.}
-        $r \leftarrow r+1$ \;
-        $s \leftarrow \min (s+1,S)$ \;
-        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
-      }
-    }
-    \If{r=R}
-    {
       Discard packet\;
@@
       Attempt transmission of 1 packet \;
+      \If{collision}{
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$ \;
+        $s \leftarrow \min (s+1,S)$ \;
+        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
+      }
+    }
     \eIf{success}{
@@
       Attempt transmission of $2^s$ packets \;
+      \If{collision}{
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$ \;
+        $s \leftarrow \min (s+1,S)$ \;
+        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
+      }
+    }
     \eIf{success}{
@@
 Hysteresis is included in \ref{alg:CSMA_ECA_hist}.
-Adding hysteresis is as simple as removing lines 13 and 23 from Algorithm \ref{alg:CSMA_ECA}.
+Adding hysteresis is as simple as removing lines 18 and 23 from Algorithm \ref{alg:CSMA_ECA}.
 Note that when hysteresis is used, the backoff stage is reset only when the node has no packet to serve.

### 2013-01-25 paper.tex
fixed spaces
Boris and Azade dropbox version

@@
-%% bare_conf.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
 \documentclass[conference]{IEEEtran}
-% Add the compsoc option for Computer Society conferences.
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[conference]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
 \usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
 % *** GRAPHICS RELATED PACKAGES ***
@@
 \fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at: 
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
-%\usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-%\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-%\usepackage[linesnumbered,noend]{algorithm2e}
 \usepackage[linesnumbered,vlined]{algorithm2e}
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
+\usepackage[ansinew]{inputenc}
+\usepackage{epsfig}
+\usepackage{amsfonts}
+\usepackage{amssymb}
+\usepackage{subfigure}
+\usepackage[usenames]{color}
+\usepackage{soul}
+\usepackage{comment}
+\usepackage{multirow}
+\usepackage[cmex10]{amsmath}
+\usepackage{bm}
+\usepackage{setspace}
 
@@
 
+\newcommand{\Az}[1]{{\bf{#1}}}
+
 \begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-\title{Evolution of CSMA Protocols for the IEEE 802.11 Standard}
+\title{Future Evolution of CSMA Protocols for the IEEE 802.11 Standard}
 
@@
 }
-%\and
-%\IEEEauthorblockN{Homer Simpson}
-%\IEEEauthorblockA{Twentieth Century Fox\\
-%Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\and
-%\IEEEauthorblockN{James Kirk\\ and Montgomery Scott}
-%\IEEEauthorblockA{Starfleet Academy\\
-%San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212\\
-%Fax: (888) 555--1212}}
-
-% conference papers do not typically use \thanks and this command
-% is locked out in conference mode. If really needed, such as for
-% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
-% after \documentclass
-
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-% 
-%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
-%Homer Simpson\IEEEauthorrefmark{2},
-%James Kirk\IEEEauthorrefmark{3}, 
-%Montgomery Scott\IEEEauthorrefmark{3} and
-%Eldon Tyrell\IEEEauthorrefmark{4}}
-%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
-%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
 
@@
 %\boldmath
-In this paper we present the requirements of candidate protocols to replace the pervasive CSMA/CA medium access control.
+In this paper we present the requirements of candidate protocols to replace the prevalent CSMA/CA medium access control.
 We discuss the possibility of further preventing collisions and provide an overview of the related work.
@@
 \end{abstract}
-% IEEEtran.cls defaults to using nonbold math in the Abstract.
-% This preserves the distinction between vectors and scalars. However,
-% if the conference you are submitting to favors bold math in the abstract,
-% then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals/conferences frown on
-% math in the abstract anyway.
-
-% no keywords
 
@@
 % no \IEEEPARstart
-This demo file is intended to serve as a ``starter file''
-for IEEE conference papers produced under \LaTeX\ using
-IEEEtran.cls version 1.7 and later.
 % You must have at least 2 lines in the paragraph with the drop letter
 % (should never be an issue)
-I wish you the best of success.
 
-A candidate to replace CSMA/CA should 
+Since the popularization of the IEEE 802.11 standard, several works \cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
+barcelo2008lba,bellalta2009vtc,he2009srb,barcelo2010fcc,fang2011dlm,hui2011epp,barcelo2011tcf} \Az{(missing any?)} have proposed modifications to CSMA/CA (Carrier Sense Multiple Access with Collision Avoidance), the contention protocol used in the standard for sharing the medium.
+\Az{Despite the throughput improvement offered by all these works, none of the proposed modifications have yet been adapted by the standard.}
+
+It is essential for a candidate CSMA/CA replacement to
 \begin{itemize}
-  \item Provide performance advantages, either in the form of throughput or short term fairness.
-  \item Be backward compatible with current implementation.
-  \item Support a large number of simultaneous contenders.
-  \item Be simple a simple evolution implementation to ease the transition and reduce time to market (Optional but desirable).
+  \item provide performance advantages in terms of throughput and/or short term fairness;
+  \item be backward compatible with the current implementation;
+  \item support a large number of simultaneous contenders; and 
+  \item be a simple evolution, in terms of implementation, to ease the transition and reduce the time to market (optional but desirable).
 \end{itemize}
-\section{Related Work}
-Since the popularization of IEEE 802.11, several papers have proposed modifications to the contention protocol that is used for sharing the medium.
-They can be categorized in three groups regarding the approach they use.
-The first one prevents that the contention window is reset to its minimum value after successful transmissions.
-Examples of this first group include \cite{bharghavan1994map,wang2004ncr}.
-This solution improves throughput in saturation conditions at the price of lowering short term fairness.
 
-The second groups involves the accurate estimation of the number of contenders to adjust the contention parameters.
-Two examples of this group are \cite{cali2000dti,lopez-toledo2006aoi}.
-This approach offers some throughput and fairness gains at the expense of increased implementation complexity.
-As the number of contenders is estimated relying on the number of collisions, the presence of channel errors further complicates the estimation.
-Furthermore, there is a fundamental trade-off between the accuracy and the reaction time of the estimation.
+The aforementioned works can be categorized in three groups regarding the approach they use.
+\Az{In this paper, we will focus on the last of these three groups, proposing an additional enhancement to make it more adaptable to network size variations, and make the case for its adaptation by the standard.}
 
-The aforementioned solutions are not able to fairly share the medium with legacy devices.
-In fact, these proposals are, generally speaking, less aggressive than the currently implemented protocol.
-Consequently, in a hypothetical mixed network in which the new and old protocols coexist, the new stations would receive a smaller share of the available bandwidth in a scenario.
+In the first group, represented by \cite{bharghavan1994map,wang2004ncr}, a throughput increase is obtained in saturation conditions by preventing the contention window from resetting to its minimum value after each successful transmission.
+However this performance gain comes at the expense of reduced short term fairness, since nodes that have recently gone through a sequence of consecutive collisions are forced to stay at a higher back-off stage and thus are further penalized by less frequent transmissions.
 
-A more important limitation of the solutions exposed so far is that the throughput is bounded by that of CSMA/CA with optimal configuration \cite{cali2000dti,bianchi2000pai}.
-In the present paper we focus on a third group of solutions that delivers throughput above the maximum attainable by CSMA/CA.
+In the second group of works, exemplified here by \cite{cali2000dti,lopez-toledo2006aoi}, an accurate estimate of the number of contenders is used to adjust the contention parameters.
+This approach offers some throughput and fairness gains, but at the expense of increased implementation complexity.
+In addition, as the number of contenders is estimated relying on the number of collisions, the presence of channel errors further complicates the estimation \Az{[Complicates, or renders it inaccurate?
+Is there a work-around offered?]}.
+Furthermore, there is a fundamental trade-off between the stability and the speed of reaction of the estimation.
 
-This third group of solutions uses a deterministic backoff after successful transmissions to further reduce the chances of collisions.
-Under certain conditions, collision-free operation is reached.
-It was introduced in \cite{barcelo2008lba} and a more detailed analysis that includes both saturated and non-saturated conditions is presented in \cite{bellalta2009vtc}.
-A more in-depth study is carried out in \cite{he2009srb}, including realistic elements such as clear channel assessment errors.
+When it comes to backward compatibility, neither of the aforementioned solutions are able to fairly share the medium with legacy devices.
+In fact, these proposals are, generally speaking, less aggressive than the currently implemented CSMA/CA protocol.
+Consequently, in a hypothetical mixed network in which the new and old protocols coexist, the new stations would receive a smaller share of the available bandwidth.
+
+A more important limitation of the two approaches exposed so far is that the throughput is bounded by that of CSMA/CA with optimal configuration \cite{cali2000dti,bianchi2000pai}.
+
+The third group of solutions, which is the focus of the current paper, can deliver a throughput above the maximum attainable by CSMA/CA.
+This performance boost is achieved mainly by the use of deterministic backoff after successful transmissions, which reduces the chances of collisions for nodes that were successful in their previous transmission.
+Furthermore, under certain conditions, a collision-free operation can be reached.
+\Az{We will refer hereafter to the class of algorithms that use deterministic backoffs after successful transmissions as CSMA with Enhanced Collision Avoidance (CSMA/ECA).} This approach was first introduced in \cite{barcelo2008lba} \Az{(Does this contain analysis?)}, and later a more detailed analysis for both saturated and non-saturated conditions was presented in \cite{bellalta2009vtc}.
+A more in-depth study is carried out in \cite{he2009srb}, including realistic elements such as the possibility of Clear Channel Assessment errors.
 Different aspects of fairness are addressed in \cite{he2009srb,barcelo2010fcc,fang2011dlm}.
-The performance in realistic channels taking into account the Auto Rate Fallback mechanism is evaluated \cite{martorell2012pec,martorell2012tfl}.
+The performance in realistic channels, taking into account the Auto Rate Fallback mechanism, is evaluated in \cite{martorell2012pec,martorell2012tfl}.
 
@@
 
-In the present paper, we study a completely distributed solution to accommodate a large number of contenders in a fair collision-free fashion.
+In the current paper, we present a completely distributed solution, called {\it backoff stage hysteresis} or simply hysteresis, to accommodate a large number of contenders in a fair collision-free fashion.
+We furthermore detail pseudo-code algorithms for three different types of CSMA/ECA and highlight how they can be easily implemented using simple modifications to the currently-used CSMA/CA algorithm.
+We will then present some performance evaluation, quantifying the performance gain that these three types of CSMA/ECA can achieve, both in terms of throughput and fairness.
+
+\Az{-comment about backward compatibility and other items in the candidate requirements.}
 
@@
   %\tcc{Initialize retransmission attempts.}
-  $r \leftarrow 0$ \;
-  %\tcc{Initialize backoff stage.}
-  $s \leftarrow 0$ \;
+  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
   %\tcc{Initialize backoff counter.}
@@
 {
-  %\tcc{Initialize retransmission attempts.}
-  $r \leftarrow 0$ \;
-  %\tcc{Initialize backoff stage.}
-  $s \leftarrow 0$ \;
-  %\tcc{Initialize backoff counter.}
+  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
   $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
@@
 }
+\vspace{0.2cm}
 \caption{CSMA/ECA with hysteresis}
@@
 }
+\vspace{0.2cm}
 \caption{CSMA/ECA with hysteresis and fair-share}
@@
 
-\section{Performance Evaluation}
+\section{Performance Evaluation} \label{sec:perf_eval}
 
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
+%Ok, I'd use CWmin/2 instead of Nmax, and "s" instead of "l" for the backoff stage.
 
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
+In this section, the performance of the different CSMA protocols is evaluated through simulations.
+The presented results show the saturation throughput and fairness (Jain's Fairness Index (JFI) \cite{jain1984quantitative}) when the number of contenders increases.
+Additionally, the evolution of the probabilities that one slot is empty, contains a successful transmission or a collision, are plotted to provide further insights on the operation of each protocol.
 
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
+To evaluate the four protocols, a network of N nodes is considered, where each node is within the coverage area of the others.
+The nodes are set to be under saturation (always have packets to transmit).
+The number of contenders, $N$, ranges from $2$ to $50$, and one thousand instances of the simulation are performed for each $N$.
+All plotted results show the 95\% confidence intervals.
 
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
+A simulator of such scenario has been built, from scratch, using the C++ language and based on the COST (Component Oriented Simulation Toolkit) libraries \cite{yucesan2002cost}.
+The specific parameters of the IEEE 802.11n amendment \cite{IEEE80211n} are considered, and listed on Table \ref{Tbl:parameters}.
+In case packets are aggregated, default A-MPDU aggregation is considered.
+Further MAC-related parameters, as well as the code for the four CSMA protocols, are available online \cite{SanabriaSimulatorECA2012}.
 
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals/conferences use top floats
-% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
-% command of the stfloats package.
+\begin{table}[h!!!]
+\centering
+\begin{tabular}{c|c}
+Parameter & Value \\
+\hline
+$CW_{\min}$ & $16$ slots \\
+$S$ & 5 stages\\
+SIFS, DIFS & $9~\mu$s, $34~\mu$s \\
+$\sigma$ & $16~\mu$s \\
+$T_{\text{PHY}}$ & $32~\mu$s \\
+MAC header length (MH) & $288$ bits \\
+Service Field length (SF) & $16$ bits \\
+MPDU Delimiter length (MD) & $32$ bits \\
+Block ACK length ($L_{\text{BACK}}$) & $256$ bits \\
+Data Length $L_{\text{DATA}}$ & 12000 bits \\
+Tail Bits (TB) & $6$ bits \\
+Data Bits Per Symbol $L_{\text{DBPS}}$ & $260$ bits/OFDM symbol \\
+OFDM Symbol Duration $T_{\text{s}}$ & $4~\mu$s \\
+\hline
+\end{tabular}
+
+\caption{Parameters considered for the performance evaluation}\label{Tbl:parameters}
+
+\end{table}
+
+%A successful slot has a duration (in $\mu$s) of
+%\begin{small}
+%\begin{align}
+% &T_{\text{slot,s}}(m)=\nonumber\\
+% &T_{\text{PHY}} + \left \lceil \frac{\text{SF} + m(\text{MH} + L_{\text{DATA}}) + (m-1)\text{MD} + \text{TB} }{L_{\text{DBPS}}} \right \rceil T_{\text{s}} +\\
+% &\text{SIFS}+ T_{\text{PHY}} + \left \lceil \frac{\text{SF} + L_{\text{BACK}} + \text{TB} }{L_{\text{DBPS}}} \right \rceil T_{\text{s}} +\text{DIFS}+\sigma \nonumber
+%\end{align}
+%\end{small}
+
+In Figure \ref{Fig:throughput}, the throughput of all four schemes is plotted.
+As can be observed, the throughput for CSMA/CA decreases with the number of nodes, as the binary exponential backoff reduces the number of transmission attempts to keep the collisions low.
+The basic CSMA/ECA is able to achieve the collision-free operation if the number of nodes is lower than the deterministic backoff value, i.e., $N\leq \text{CW}_{\min}/2$.
+This is why in this figure, a phase transition is observed at $N=8$.
+When $N$ is larger than $\text{CW}_{\min}/2=8$, although the collision-free operation is not possible, the throughput obtained remains higher than in CSMA/CA.
+This is because, regardless of whether or not the collision-free operation is reached, in CSMA/ECA, a node always has a lower collision probability immediately after a successful transmission, without affecting the collision probability of other nodes.
+
+\begin{figure}[ht!!!!!!!!]
+\centering
+\epsfig{file=throughput-combined.eps,scale=0.45,angle=270}\label{Fig:throughput}
+\caption{Throughput}
+\end{figure}
+
+\begin{figure}[ht!!!!!!!!]
+\centering
+\epsfig{file=fairness-combined.eps,scale=0.45,angle=270}\label{Fig:fairness}
+\caption{Fairness}
+\end{figure}
+
+\begin{figure}[ht!!!!!!!!]
+\centering
+\epsfig{file=avgCollisions-6sta.eps,scale=0.45,angle=270}\label{Fig:Pc_6}
+\caption{Cumulative fraction of slots spent in collision for $N=6$ nodes}
+\end{figure}
+
+\begin{figure}[ht!!!!!!!!]
+\centering
+\epsfig{file=avgCollisions-12sta.eps,scale=0.45,angle=270}\label{Fig:Pc_12}
+\caption{Cumulative fraction of slots spent in collision for $N=12$ nodes}
+\end{figure}
+
+Clearly, reaching a collision-free operation is highly desirable, but not always possible in the basic CSMA/ECA.
+CSMA/ECA with Hysteresis exactly addresses this issue by simply not resetting the backoff stage after a successful transmission.
+In other words, the deterministic backoff value ($2^s \text{CW}_{\min}/2$) is chosen based on the backoff stage, $s$, in which the transmission was successful.
+This way, a larger number of contenders can reach a collision-free operation (Figure \ref{Fig:throughput}), but at the expense of a lower throughput for $N\leq 8$ compared to the basic CSMA/ECA.
+This is because if collision-free operation can be reached with a lower deterministic backoff value, there will be fewer empty slots, hence higher channel access efficiency.
+Furthermore, CSMA/ECA with hysteresis has a lower long-term fairness (Figure \ref{Fig:fairness}) compared to the basic CSMA/ECA or even the legacy CSMA/CA protocol, as once the network has reached the collision-free state, nodes that use a large backoff stage will have fewer transmission opportunities than others.
+
+To address the aforementioned fairness issue when hysteresis is used, CSMA/ECA with hysteresis and fair-share was introduced.
+It achieves greater throughput than CSMA/ECA with hysteresis only, while being fair for any number of contenders.
+This is achieved by using packet aggregation, transmitting $2^s$ packets per attempt.
+As observed in Figure \ref{Fig:throughput}, using aggregation we are not only able to provide a fair access, but also we are able to significantly increase the throughput.
+\Az{Note that the throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network in terms of channel use.}
+
+In Figures \ref{Fig:Pc_6} and \ref{Fig:Pc_12}, the temporal (in number of slots) evolution of the cumulative fraction of slots spent in collision is shown for $N = 6$ and $12$ nodes.
+As expected, for $N=6$, with all CSMA/ECA variants, collision-free operation is achieved (since $N<8$) and the probability that a slot contains a collision tends rapidly to zero.
+The collision-free operation is reached faster when hysteresis is used.
+When $N=12$, only the protocols using hysteresis reach collision-free operation.
 
@@
 
-The authors would like to thank...
+The authors would like to thank Boris for buying a fan, which can be useful to make fun.
 

### 2013-01-28 paper.tex
Azadeh submitted version

@@
 %\boldmath
-In this paper we present the requirements of candidate protocols to replace the prevalent CSMA/CA medium access control.
-We discuss the possibility of further preventing collisions and provide an overview of the related work.
-We specify a protocol that is a candidate for replacing CSMA/CA in pseudocode and use simulation to assess performance metrics such as throughput, fairness and collision probability.
+In this paper a candidate protocol to replace the prevalent CSMA/CA medium access control in Wireless Local Area Networks is presented.
+The proposed protocol can achieve higher throughput compared to CSMA/CA, while maintaining fairness, and without additional implementation complexity.
+Under certain circumstances, it is able to reach and maintain collision-free operation, even when the number of contenders is variable and potentially large.
+It is backward compatible, allowing for new and legacy stations to coexist without degrading one another's performance, a property that can make the adoption process by future versions of the standard smooth and inexpensive.
+
+% In this paper we have described a simply implementable protocol that can be used as a replacement of the prevalent CSMA/CA protocol used in the IEEE 802.11 standard.
+Compared to the CSMA/CA protocol, the proposed protocol, called CSMA/ECA with hysteresis and fair-share, offers performance boost in terms of throughput while preserving long-term fairness.
+In fact, under certain conditions, collision-free operation can be reached in a distributed and adaptive manner, even as the number of nodes in the network increase.
+Furthermore, this protocol is backward compatible, and can operate with nodes using the legacy protocol without degrading the performance of the legacy nodes.
+All these properties make the proposed protocol a very suitable candidate to replace CSMA/CA in the upcoming revisions of the standard.
+
+% In this paper we present the requirements of candidate protocols to replace the prevalent CSMA/CA medium access control.
+% We discuss the possibility of further preventing collisions and provide an overview of the related work.
+% We specify a protocol that is a candidate for replacing CSMA/CA in pseudo-code and use simulation to assess performance metrics such as throughput, fairness and collision probability.
 \end{abstract}
@@
 
-Since the popularization of the IEEE 802.11 standard, several works \cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
-barcelo2008lba,bellalta2009vtc,he2009srb,barcelo2010fcc,fang2011dlm,hui2011epp,barcelo2011tcf} \Az{(missing any?)} have proposed modifications to CSMA/CA (Carrier Sense Multiple Access with Collision Avoidance), the contention protocol used in the standard for sharing the medium.
-\Az{Despite the throughput improvement offered by all these works, none of the proposed modifications have yet been adapted by the standard.}
+Since the popularization of the IEEE 802.11 standard, several works (e.g.,  \cite{bharghavan1994map,wang2004ncr,cali2000dti,lopez-toledo2006aoi,
+barcelo2008lba,bellalta2009vtc,he2009srb,barcelo2010fcc,fang2011dlm,hui2011epp,barcelo2011tcf}) have proposed modifications to CSMA/CA (Carrier Sense Multiple Access with Collision Avoidance), the contention protocol used in the standard for controling the access to the shared medium.
+{Despite the throughput improvement offered by all these works, none of the proposed modifications have yet been adopted by the standard.}
 
-It is essential for a candidate CSMA/CA replacement to
+A suitable candidate for the replacement of CSMA/CA, should
 \begin{itemize}
-  \item provide performance advantages in terms of throughput and/or short term fairness;
+  \item provide performance advantages in terms of throughput and/or short-term fairness;
   \item be backward compatible with the current implementation;
@@
 The aforementioned works can be categorized in three groups regarding the approach they use.
-\Az{In this paper, we will focus on the last of these three groups, proposing an additional enhancement to make it more adaptable to network size variations, and make the case for its adaptation by the standard.}
+{In this paper, we will focus on the last of these three groups, proposing an additional enhancement to make it more adaptable to network size variations, and make the case for its adoption by the future versions of the standard.}
 
-In the first group, represented by \cite{bharghavan1994map,wang2004ncr}, a throughput increase is obtained in saturation conditions by preventing the contention window from resetting to its minimum value after each successful transmission.
-However this performance gain comes at the expense of reduced short term fairness, since nodes that have recently gone through a sequence of consecutive collisions are forced to stay at a higher back-off stage and thus are further penalized by less frequent transmissions.
+In the first group, represented here by \cite{bharghavan1994map,wang2004ncr}, a throughput increase is obtained in saturation conditions by preventing the contention window from resetting to its minimum value after each successful transmission.
+However this performance gain comes at the expense of reduced short-term fairness, since nodes that have recently gone through a sequence of consecutive collisions are forced to stay at a higher backoff stage and thus are further penalized by less frequent transmissions.
 
@@
 This approach offers some throughput and fairness gains, but at the expense of increased implementation complexity.
-In addition, as the number of contenders is estimated relying on the number of collisions, the presence of channel errors further complicates the estimation \Az{[Complicates, or renders it inaccurate?
-Is there a work-around offered?]}.
+In addition, as the number of contenders is estimated relying on the number of collisions, the presence of channel errors renders the estimation inaccurate.
 Furthermore, there is a fundamental trade-off between the stability and the speed of reaction of the estimation.
@@
 Furthermore, under certain conditions, a collision-free operation can be reached.
-\Az{We will refer hereafter to the class of algorithms that use deterministic backoffs after successful transmissions as CSMA with Enhanced Collision Avoidance (CSMA/ECA).} This approach was first introduced in \cite{barcelo2008lba} \Az{(Does this contain analysis?)}, and later a more detailed analysis for both saturated and non-saturated conditions was presented in \cite{bellalta2009vtc}.
+{We will refer hereafter to the class of algorithms that use deterministic backoffs after successful transmissions as CSMA with Enhanced Collision Avoidance (CSMA/ECA).} This approach was first introduced in \cite{barcelo2008lba}, and later a more detailed analysis for both saturated and non-saturated conditions was presented in \cite{bellalta2009vtc}.
 A more in-depth study is carried out in \cite{he2009srb}, including realistic elements such as the possibility of Clear Channel Assessment errors.
@@
 
-Even though initial research efforts where focused on the WLAN collision problem, more recent papers try to extend the idea to multi-hop networks.
-In \cite{hui2011epp}, the multi-hop slotted case explored.
-The more realistic situation in multi-hop networks in which the time is not slotted is studied in \cite{barcelo2013dcc}.
+Even though initial research efforts where focused on the WLAN collision problem, more recent works try to extend the idea to multi-hop networks.
+In \cite{hui2011epp}, the multi-hop slotted case is explored.
+A solution that does not require network-wide slot synchronization is studied in  \cite{barcelo2013dcc}.
 
@@
 
-In the current paper, we present a completely distributed solution, called {\it backoff stage hysteresis} or simply hysteresis, to accommodate a large number of contenders in a fair collision-free fashion.
-We furthermore detail pseudo-code algorithms for three different types of CSMA/ECA and highlight how they can be easily implemented using simple modifications to the currently-used CSMA/CA algorithm.
+In the current paper, we present a completely distributed solution, called {\it backoff stage hysteresis} or simply hysteresis, which is a simple modification to CSMA/ECA to accommodate a large number of contenders in an adaptive and distributed fashion.
+Hysteresis by itself, comes at the expense of reduction in long-term fairness.
+Using hysteresis together with what we will refer to as {\it fare-share} allows for reaching collision-free operation with self-adaptation to a variable, and potentially large, number of contenders while preserving fairness.
+The concept of fair-share was first introduced in \cite{fang2011dlm}.
+We will furthermore detail pseudo-code algorithms for three different types of CSMA/ECA (basic, with hysteresis, and with both hysteresis and fair-share) and highlight how they can be easily implemented using simple modifications to the currently-used CSMA/CA algorithm.
 We will then present some performance evaluation, quantifying the performance gain that these three types of CSMA/ECA can achieve, both in terms of throughput and fairness.
 
-\Az{-comment about backward compatibility and other items in the candidate requirements.}
+{Note that the coexistence and fairness of CSMA/ECA and CSMA/CA has been studied in \cite{barcelo2010fcc}.
+The addition of hysteresis and fair-share may represent less frequent transmissions and more packets per transmission, which do not interfere in any way with the execution of the legacy protocol.}
 
@@
 
-\begin{algorithm}
+In this section and the next, we make the case for the suitability of CSMA/ECA with hysteresis and fair-share as a future replacement to the CSMA/CA protocol.
+To do this, we begin by presenting a simple pseudo-code which captures the essence of the CSMA/CA protocol as it is currently implemented in the IEEE 802.11 standard.
+We will then show how CSMA/ECA can be implemented by a minor modification to the CSMA/CA algorithm.
+We then describe hysteresis and fair-share, by presenting minor modifications, one by one, to the CSMA/ECA algorithm.
+The final product---CSMA/ECA with hysteresis and fair-share---satisfies the four requirements specified in the introduction, which makes it a suitable candidate to replace CSMA/CA in the upcoming revisions of the standard.
+We would like to emphasize here that the basic CSMA/ECA and the concept of fair-share are not presented here for the first time, however, we include the algorithms corresponding to these protocols for both completeness and ease of comparison between protocols.
+
+Algorithm \ref{alg:CSMA_CA} describes the CSMA/CA algorithm that is used in current networks.
+When a station joins the contention, it initializes the retry attempt counter $r$ and the backoff stage $s$ to zero.
+The backoff counter $b$ is initialized using a uniform random distribution and the minimum contention window $\text{CW}_{\min}$.
+After each collision, the retry attempt counter and the backoff stage counter are incremented.
+As a consequence of the incremented backoff stage, a larger contention window is used.
+Note that there is a maximum backoff stage $S$ and a maximum retry limit $R$ specified by the protocol.
+When the number of transmission attempts on a packet reaches the maximum retry limit, the packet is discarded.
+Furthermore, $r$ and $s$ are reset to zero, and a new value for $b$ is computed both when the retry limit is reached and when the transmission is successful.
+Note that the pseudo-code performs the same action in lines $18$ and $21$.
+Therefore, the algorithm could have been further simplified if instead this action were moved to immediately after line $16$.
+However, the current presentation of the algorithm eases the comparison between the CSMA/CA and the enhancements that follow.
+
+% In this section, we describe the original CSMA/CA protocol and then we describe each of the modifications that we propose, one by one.
+% 
+% 
+% In this section we describe the CSMA/CA protocol as it is currently implemented in the IEEE 802.11 and an evolution of the standard that satisfies the four requirements specified in the introduction.
+% Therefore, the presented protocol is a valid candidate to replace CSMA/CA in the upcoming revisions of the standard.
+
+There are three changes in the CSMA/ECA with hysteresis and fair-share compared to the legacy CSMA/CA protocol.
+Firstly, a deterministic backoff is used after successful transmissions.
+This simple modification converts CSMA/CA into CSMA/ECA.
+Secondly, the backoff stage is not reset after a packet is serviced.
+The backoff stage is reset only when the station leaves the contention because it has no packet to be transmitted. This modification is what we have been referring to as hysteresis.
+And thirdly, the number of packets transmitted in every transmission attempt is chosen as a function of the backoff stage, which is what we call fair-share.
+Note that current standards already support the transmission of multiple packets in a single slot.
+
+Algorithm \ref{alg:CSMA_ECA} describes the basic CSMA/ECA in which a deterministic backoff is used after successful transmissions.
+The only change with respect to CSMA/CA is in fact in line $14$, where the random assignment of $b$ in CSMA/CA is replaced by a deterministic assignment in CSMA/ECA.
+Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2$.
+
+Hysteresis is included in Algorithm \ref{alg:CSMA_ECA_hist}.
+Adding hysteresis is as simple as removing line $16$ from Algorithm \ref{alg:CSMA_ECA}.
+This means that the deterministic backoff value chosen after a successful transmission in this case is half the length of the backoff stage in which the successful transmission has occurred.
+In fact, the backoff stage is reset only when the node has no packet to serve.
+This allows for the use of larger deterministic backoff values, which means that the network can reach a collision-free operation for a higher number of nodes.
+Since in the collision-free operation, some nodes may have higher backoff values than others, this protocol does not share the medium in a fair manner among nodes.
+
+Finally, fair-share is implemented in Algorithm \ref{alg:CSMA_ECA_hist_fair} by simply modifying
+line $9$ of Algorithm \ref{alg:CSMA_ECA_hist} to increase the number of transmitted packets from one to $2^s$.
+This way, at each transmission opportunity in collision-free operation, each node transmits a number of packets proportional to the length of its deterministic backoff.
+Therefore, nodes that have to wait longer between transmission opportunities get to transmit more packets, hence fairness is preserved.
+
+\begin{algorithm}[ht!!!]
 \While{the device is on}
@@
     }
+    $r \leftarrow 0$ \;
+    $s \leftarrow 0$ \;
 	 \eIf{success}{
       %\tcc{Random backoff.}
-      $r \leftarrow 0$ \;
-      $s \leftarrow 0$ \;
       $b \leftarrow \mathcal{U}[0,2^{s}\rm{CW}_{min}-1]$\;
@@
       Discard packet\;
-      $r \leftarrow 0$ \;
-      $s \leftarrow 0$ \;
       $b \leftarrow \mathcal{U}[0,2^s \rm{CW}_{min}-1]$\;
@@
 
-\begin{algorithm}
+\begin{algorithm}[ht!!!]
 \While{the device is on}
@@
     }
+    $r \leftarrow 0$ \;
+    $s \leftarrow 0$ \;    
     \eIf{success}{
       %\tcc{Random backoff.}
-      $r \leftarrow 0$ \;
-      $s \leftarrow 0$ \;
       $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
@@
       Discard packet\;
-      $r \leftarrow 0$ \;
-      $s \leftarrow 0$ \;
       $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
@@
     }
+    $r \leftarrow 0$ \;
+    %$s \leftarrow 0$ \;
     \eIf{success}{
       %\tcc{Random backoff.}
-      $r \leftarrow 0$ \;
-      %$s \leftarrow 0$ \;
       $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
@@
       Discard packet\;
-      $r \leftarrow 0$ \;
-      %$s \leftarrow 0$ \;
       $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
@@
 
-In this section we describe the CSMA/CA protocol as it is currently implemented in the IEEE 802.11 and an evolution of the standard that satisfies the four requirements specified in the introduction.
-Therefore, the presented protocol is a valid candidate to replace CSMA/CA in the upcoming revisions of the standard.
-
-There are three changes in the new protocol compared to the legacy one.
-Firstly, a deterministic backoff is used after successful transmissions.
-Secondly, the backoff stage is not reset after a packet is serviced.
-The backoff stage is reset only when the station leaves the contention because it has no packet to be transmitted.
-And thirdly, the number of packets transmitted in every transmission attempt is a function of the backoff stage.
-Note that current standards already support the transmission of multiple packets in a single slot.
-
-In the following we describe the original protocol and then we describe each of the modifications that we propose, one by one.
-
-Algorithm \ref{alg:CSMA_CA} describes the CSMA/CA algorithm that is used in current networks.
-When a station joins the contention it initializes the retry attempt counter $r$ and the backoff stage $s$ to zero.
-The backoff counter $b$ is initialized using a uniform random distribution and a contention window.
-After each collision, the retry attempt counter and the backoff stage counter are incremented.
-As a consequence of the incremented backoff stage, a larger contention window is used.
-Note that there is maximum backoff stage $S$ and a maximum retry limit $R$.
-When the maximum retry  limit is reached the packet is discarded, $r$ and $s$ are reset anb a new value for $b$ is computed.
-Similarly, if the packet is successfully transmitted, the $s$ and $r$ counters are reset and a new random backoff counter is computed.
-
-Algorithm \ref{alg:CSMA_ECA} describes CSMA/ECA in which a deterministic backoff is used after successful transmission.
-The only change with respect to CSMA/CA is in line 14, where the determimistic assignment takes place.
-
-Hysteresis is included in \ref{alg:CSMA_ECA_hist}.
-Adding hysteresis is as simple as removing lines 18 and 23 from Algorithm \ref{alg:CSMA_ECA}.
-Note that when hysteresis is used, the backoff stage is reset only when the node has no packet to serve.
-
-Finally, fair-share is implemented in \ref{alg:CSMA_ECA_hist_fair}.
-Line 10 of \ref{alg:CSMA_ECA_hist} is modified to increase the number of transmitted packets from one to $2^s$.
-
 \section{Performance Evaluation} \label{sec:perf_eval}
@@
 The presented results show the saturation throughput and fairness (Jain's Fairness Index (JFI) \cite{jain1984quantitative}) when the number of contenders increases.
-Additionally, the evolution of the probabilities that one slot is empty, contains a successful transmission or a collision, are plotted to provide further insights on the operation of each protocol.
+Additionally, the evolution of the probabilities that one slot collision is plotted to provide further insights on the operation of each protocol, and specially whether the collision-free operation is reached.
 
-To evaluate the four protocols, a network of N nodes is considered, where each node is within the coverage area of the others.
+%Additionally, the evolution of the probabilities that one slot is empty, contains a successful transmission or a collision, are plotted to provide further insights on the operation of each protocol.
+
+To evaluate the four protocols, a network of $N$ nodes is considered, where each node is within the coverage area of the others.
 The nodes are set to be under saturation (always have packets to transmit).
@@
 A simulator of such scenario has been built, from scratch, using the C++ language and based on the COST (Component Oriented Simulation Toolkit) libraries \cite{yucesan2002cost}.
-The specific parameters of the IEEE 802.11n amendment \cite{IEEE80211n} are considered, and listed on Table \ref{Tbl:parameters}.
+The specific parameters of the IEEE 802.11n amendment \cite{IEEE80211n} are considered, and listed in Table \ref{Tbl:parameters}.
 In case packets are aggregated, default A-MPDU aggregation is considered.
@@
 $S$ & 5 stages\\
-SIFS, DIFS & $9~\mu$s, $34~\mu$s \\
-$\sigma$ & $16~\mu$s \\
-$T_{\text{PHY}}$ & $32~\mu$s \\
-MAC header length (MH) & $288$ bits \\
-Service Field length (SF) & $16$ bits \\
-MPDU Delimiter length (MD) & $32$ bits \\
-Block ACK length ($L_{\text{BACK}}$) & $256$ bits \\
-Data Length $L_{\text{DATA}}$ & 12000 bits \\
-Tail Bits (TB) & $6$ bits \\
-Data Bits Per Symbol $L_{\text{DBPS}}$ & $260$ bits/OFDM symbol \\
-OFDM Symbol Duration $T_{\text{s}}$ & $4~\mu$s \\
+% SIFS, DIFS & $9~\mu$s, $34~\mu$s \\
+% $\sigma$ & $16~\mu$s \\
+% $T_{\text{PHY}}$ & $32~\mu$s \\
+% MAC header length (MH) & $288$ bits \\
+% Service Field length (SF) & $16$ bits \\
+% MPDU Delimiter length (MD) & $32$ bits \\
+% Block ACK length ($L_{\text{BACK}}$) & $256$ bits \\
+Data Packet Length & 12000 bits \\
+Data Rate & $65$ Mbps\\
+% Tail Bits (TB) & $6$ bits \\
+% Data Bits Per Symbol $L_{\text{DBPS}}$ & $260$ bits/OFDM symbol \\
+% OFDM Symbol Duration $T_{\text{s}}$ & $4~\mu$s \\
 \hline
@@
 \centering
-\epsfig{file=throughput-combined.eps,scale=0.45,angle=270}\label{Fig:throughput}
-\caption{Throughput}
+\epsfig{file=throughput-combined.eps,scale=0.45,angle=270}
+\caption{Throughput}\label{Fig:throughput}
 \end{figure}
@@
 \centering
-\epsfig{file=fairness-combined.eps,scale=0.45,angle=270}\label{Fig:fairness}
-\caption{Fairness}
+\epsfig{file=fairness-combined.eps,scale=0.45,angle=270}
+\caption{Fairness}\label{Fig:fairness}
 \end{figure}
@@
 \centering
-\epsfig{file=avgCollisions-6sta.eps,scale=0.45,angle=270}\label{Fig:Pc_6}
-\caption{Cumulative fraction of slots spent in collision for $N=6$ nodes}
+\epsfig{file=avgCollisions-6sta.eps,scale=0.45,angle=270}
+\caption{Cumulative fraction of slots spent in collision for $N=6$ nodes}\label{Fig:Pc_6}
 \end{figure}
@@
 \centering
-\epsfig{file=avgCollisions-12sta.eps,scale=0.45,angle=270}\label{Fig:Pc_12}
-\caption{Cumulative fraction of slots spent in collision for $N=12$ nodes}
+\epsfig{file=avgCollisions-12sta.eps,scale=0.45,angle=270}
+\caption{Cumulative fraction of slots spent in collision for $N=12$ nodes}\label{Fig:Pc_12}
 \end{figure}
@@
 As observed in Figure \ref{Fig:throughput}, using aggregation we are not only able to provide a fair access, but also we are able to significantly increase the throughput.
-\Az{Note that the throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network in terms of channel use.}
+{Note that the throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network in terms of channel use.}
 
-In Figures \ref{Fig:Pc_6} and \ref{Fig:Pc_12}, the temporal (in number of slots) evolution of the cumulative fraction of slots spent in collision is shown for $N = 6$ and $12$ nodes.
+In Figures \ref{Fig:Pc_6} and \ref{Fig:Pc_12}, the evolution of the cumulative fraction of slots spent in collision is shown for $N = 6$ and $12$ nodes.
 As expected, for $N=6$, with all CSMA/ECA variants, collision-free operation is achieved (since $N<8$) and the probability that a slot contains a collision tends rapidly to zero.
@@
 \section{Conclusion}
-The conclusion goes here.
+
+In this paper we have described a suitable replacement of the prevalent CSMA/CA protocol used in the IEEE 802.11 standard.
+Compared to the CSMA/CA protocol, the proposed protocol, called CSMA/ECA with Hysteresis and Fair-Share, offers performance boost in terms of throughput while preserving long-term fairness.
+In fact, under certain conditions, collision-free operation can be reached in a distributed and adaptive manner, even as the number of nodes in the network increases.
+Furthermore, this protocol is backward compatible, and can operate with nodes using the legacy protocol without degrading the performance of the legacy nodes.
+All these properties make the proposed protocol a good candidate to replace CSMA/CA in the upcoming revisions of the standard.
+
+% In this paper the CSMA/ECA protocol with Hysteresis and Fair Share has been proposed as a plausible replacement of the CSMA/CA protocol for IEEE 802.11 wireless local area networks.
+The protocol can achieve collision-free operation, which results in significant performance gains, with no additional complexity in term of implementation.
+In those circumstances where collision-free operation may not be maintained (e.g., due to noise), CSMA/ECA still achieves better throughput compared to CSMA/CA.
+Furthermore, it is backward compatible with the default CSMA/CA protocol.
+The Hysteresis property allows the CSMA/ECA protocol to adapt to a variable, and potentially large, number of contenders.
+However, it introduces long-term unfairness, as not all nodes in the network are able to transmit the same amount of data.
+To solve this issue, packet aggregation is used in CSMA/ECA protocol with Hysteresis and Fair Share, enabling nodes that access the channel less frequently to transmit more packets at each attempt.
+
+% 
+% \Az{The coexistence and fairness of CSMA/ECA and CSMA/CA has been studied in [8].
+The addition of hysteresis and fair-share may represent less transmissions and more packets per transmission, which do no interfere in any way with the execution of the legacy protocol.}
 
@@
 % use section* for acknowledgement
-\section*{Acknowledgment}
-
-The authors would like to thank Boris for buying a fan, which can be useful to make fun.
 

### 2013-02-28 paper.tex
working on camera ready version

@@
 \usepackage[linesnumbered,vlined]{algorithm2e}
-\usepackage[ansinew]{inputenc}
 \usepackage{epsfig}
@@
 In this paper a candidate protocol to replace the prevalent CSMA/CA medium access control in Wireless Local Area Networks is presented.
-The proposed protocol can achieve higher throughput compared to CSMA/CA, while maintaining fairness, and without additional implementation complexity.
+The proposed protocol can achieve higher throughput than CSMA/CA, while maintaining fairness, and without additional implementation complexity.
 Under certain circumstances, it is able to reach and maintain collision-free operation, even when the number of contenders is variable and potentially large.
@@
 
+Quality-of-service and traffic differentiation in CSMA/ECA has is discussed in \cite{barcelo2009tpc, he2009srb}.
+The performance of CSMA/ECA in the presence of channel errors is considered in \cite{he2009srb, barcelo2010fcc, fang2011dlm, barcelo2012mdc}. 
+Other aspects such as coexistence with legacy stations and performance in the presence of new entrants have also been covered in \cite{he2009srb, barcelo2010fcc, fang2011dlm}.
+Results for the impact of slot drift (non-ideal clocks) in protocols similar to CSMA/ECA can be found in \cite{gong2012asd}.
+In the present workshop paper, we focus on a simple greenfield scenario with a single traffic class, ideal channel, ideal clocks, and saturation conditions.
+We are currently evaluating more complex scenarios that we outline in Section \ref{sec:}
+
 Even though initial research efforts where focused on the WLAN collision problem, more recent works try to extend the idea to multi-hop networks.
@@
 
-{Note that the coexistence and fairness of CSMA/ECA and CSMA/CA has been studied in \cite{barcelo2010fcc}.
-The addition of hysteresis and fair-share may represent less frequent transmissions and more packets per transmission, which do not interfere in any way with the execution of the legacy protocol.}
+Note that the coexistence and fairness of basic CSMA/ECA and CSMA/CA has been studied in \cite{barcelo2010fcc}. 
+The addition of hysteresis and fair-share to the basic CSMA/ECA may represent less frequent transmissions and more packets per transmission
+These two modifications do not interfere in any way with the execution of the legacy protocol and, therefore, CSMA/CA with hysteresis and fair-share can also coexist with the currently deployed protocol.
 
@@
 Algorithm \ref{alg:CSMA_ECA} describes the basic CSMA/ECA in which a deterministic backoff is used after successful transmissions.
-The only change with respect to CSMA/CA is in fact in line $14$, where the random assignment of $b$ in CSMA/CA is replaced by a deterministic assignment in CSMA/ECA.
+The only change with respect to CSMA/CA is in fact in line $18$, where the random assignment of $b$ in CSMA/CA is replaced by a deterministic assignment in CSMA/ECA.
 Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2$.
+This value is roughly equal to the expectation of the random backoff used in the same situation in CSMA/CA.
+This particular choice guarantees fairness between new stations and legacy stations.
 
@@
 Adding hysteresis is as simple as removing line $16$ from Algorithm \ref{alg:CSMA_ECA}.
-This means that the deterministic backoff value chosen after a successful transmission in this case is half the length of the backoff stage in which the successful transmission has occurred.
+This means that the deterministic backoff value chosen after a successful transmission in this case is half the length of the contention window of the backoff stage in which the successful transmission has occurred.
 In fact, the backoff stage is reset only when the node has no packet to serve.
@@
 To evaluate the four protocols, a network of $N$ nodes is considered, where each node is within the coverage area of the others.
-The nodes are set to be under saturation (always have packets to transmit).
+The channel does not introduce errors and the nodes are set to be under saturation (always have packets to transmit).
 The number of contenders, $N$, ranges from $2$ to $50$, and one thousand instances of the simulation are performed for each $N$.
@@
 
-% use section* for acknowledgement
+\section*{Acknowledgment}
+
+The authors would like to thank the anonymous reviewers for their corrections, helpful comments and suggestions for future work.
+The authors are also thankful to Alessandro Checco for helping in the preparation of the related work section.
 

### 2013-02-28 paper.tex
future work

@@
 
+\section{Future Work}
+
+In this workshop paper, we have evaluated the protocols under discussion in a very simple scenario using ideal channel, ideal clocks and saturation conditions.
+Furthermore, all the participating stations use the same protocol and have a single traffic  class.
+The presented metrics include only throughput and fairness, whereas delay, jitter and packet have been omitted.
+Another scenario of interest that we have not included is the situation in which the stations join the contention to transmit a file and then they leave.
+
+Our proposed protocol is based on CSMA/ECA, which works well in non-ideal conditions, and the results we are currently obtaining confirm that CSMA/ECA with hysteresis and fair-share inherits the flexibility and robustness of its predecessors.
+
+Regarding the implementation in real prototypes, it has been proven difficult until now, as the protocols are embedded in the networks interfaces and it is not possible to reprogram them.
+There are current advances in that direction \cite{tinnirello2012wmp} but it is still difficult to set the backoff to a deterministic value after successful transmissions in available hardware.
+
+Our intention is to try our protocols in a different technology, namely radio-frequency identification (RFID), where there is programmable hardware that can be used to implement arbitrary protocols.
+RFID uses also a contention protocol which are prone to collisions, and we believe it can be a first area in which we will be available to develop working prototypes to validate some of the ideas proposed in the present paper.
+
 \section{Conclusion}

### 2013-03-01 paper.tex
playing with colors

@@
 In fact, these proposals are, generally speaking, less aggressive than the currently implemented CSMA/CA protocol.
-Consequently, in a hypothetical mixed network in which the new and old protocols coexist, the new stations would receive a smaller share of the available bandwidth.
+Consequently, in a hypothetical mixed network in which the new and old protocols coexisted, the new stations would receive a smaller share of the available bandwidth.
 
@@
 In the present workshop paper, we focus on a simple greenfield scenario with a single traffic class, ideal channel, ideal clocks, and saturation conditions.
-We are currently evaluating more complex scenarios that we outline in Section \ref{sec:}
+We are currently evaluating more complex scenarios that we outline in Section \ref{sec:future_work}
 
@@
 Hysteresis by itself, comes at the expense of reduction in long-term fairness.
-Using hysteresis together with what we will refer to as {\it fare-share} allows for reaching collision-free operation with self-adaptation to a variable, and potentially large, number of contenders while preserving fairness.
+Using hysteresis together with what we will refer to as {\it fair-share} allows for reaching collision-free operation with self-adaptation to a variable, and potentially large, number of contenders while preserving fairness.
 The concept of fair-share was first introduced in \cite{fang2011dlm}.
@@
   %\tcc{Initialize retransmission attempts.}
-  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
+  \colorbox{red}{$r \leftarrow 0$ ; $s \leftarrow 0$} \;
   %\tcc{Initialize backoff counter.}
@@
 \section{Future Work}
+\label{sec:future_work}
 

### 2013-03-01 paper.tex
working on the figure

@@
 Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2$.
-This value is roughly equal to the expectation of the random backoff used in the same situation in CSMA/CA.
+This value is roughly equal to the expectation of the backoff chosen in line $18$ in Algorithm \ref{alg:CSMA_CA}.
 This particular choice guarantees fairness between new stations and legacy stations.
@@
   %\tcc{Initialize retransmission attempts.}
-  \colorbox{red}{$r \leftarrow 0$ ; $s \leftarrow 0$} \;
+  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
   %\tcc{Initialize backoff counter.}
@@
       }
-      Attempt transmission of 1 packet \;
+      \colorbox{yellow}{Attempt transmission of 1 packet } \;
       \If{collision}{
@@
     $r \leftarrow 0$ \;
-    $s \leftarrow 0$ \;
+    \colorbox{yellow}{$s \leftarrow 0$ } \;
 	 \eIf{success}{
       %\tcc{Random backoff.}
-      $b \leftarrow \mathcal{U}[0,2^{s}\rm{CW}_{min}-1]$\;
+      \colorbox{yellow}{$b \leftarrow \mathcal{U}[0,2^{s}\rm{CW}_{min}-1]$} \;
     }
@@
       }
-      Attempt transmission of 1 packet \;
+      \colorbox{yellow}{Attempt transmission of 1 packet } \;
       \If{collision}
@@
     $r \leftarrow 0$ \;
-    $s \leftarrow 0$ \;    
+    \colorbox{yellow}{$s \leftarrow 0$} \;    
     \eIf{success}{
       %\tcc{Random backoff.}
-      $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
+      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
     }
@@
       %\tcc{Random backoff.}
-      $b \leftarrow (2^{s}\rm{CW}_{min})/2$\;
+      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
     }
@@
       }
-      Attempt transmission of $2^s$ packets \;
+      \colorbox{yellow}{Attempt transmission of $2^s$ packets } \;
       \If{collision}{
@@
 \end{algorithm}
+
+An example of the operation of CSMA/ECA with hysteresis is presented in Fig. \ref{fig:csma_eca_different_backoff}.
+
+\begin{figure*}[!t]
+\centering
+\includegraphics[width=7.0in]{figures/csma_eca_different_backoff}
+\caption{Schedule length distributed adaption example. All the stations transmit the same number of packets in each cycle, despite using different schedule lengths.}
+\label{fig:csma_eca_different_backoff}
+\end{figure*}
 
@@
 As observed in Figure \ref{Fig:throughput}, using aggregation we are not only able to provide a fair access, but also we are able to significantly increase the throughput.
-{Note that the throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network in terms of channel use.}
+{Note that the throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network.}
 
@@
 
-Regarding the implementation in real prototypes, it has been proven difficult until now, as the protocols are embedded in the networks interfaces and it is not possible to reprogram them.
+Regarding the implementation in real prototypes, it has been proven difficult until now, as the IEEE 802.11 protocols are embedded in the network interfaces and it is not possible to reprogram them.
 There are current advances in that direction \cite{tinnirello2012wmp} but it is still difficult to set the backoff to a deterministic value after successful transmissions in available hardware.
@@
 Our intention is to try our protocols in a different technology, namely radio-frequency identification (RFID), where there is programmable hardware that can be used to implement arbitrary protocols.
-RFID uses also a contention protocol which are prone to collisions, and we believe it can be a first area in which we will be available to develop working prototypes to validate some of the ideas proposed in the present paper.
+RFID uses also a contention protocol which is prone to collisions, and we believe it can be a first area in which we will be available to develop working prototypes to validate some of the ideas proposed in the present paper.
 

### 2013-03-01 paper.tex
finished figure

@@
 An example of the operation of CSMA/ECA with hysteresis is presented in Fig. \ref{fig:csma_eca_different_backoff}.
+There are four stations in the example and the transmissions of those stations are represented as rounded boxes.
+The small numbers are the value of the backoff counter of the stations, that decrement in every slot boundary.
+
+The stations initially pick a random backoff. 
+Those stations with the same backoff value collide and choose a new backoff value.
+When all the stations have a different backoff value, no more collisions will occur.
+At this point the behaviour of the system is cyclic and deterministic.
+Note that the state of the system at the beginning of the cycle is the backoff counter of the four stations: 4, 9, 14 and 2.
+It is exactly the same state that the first slot of the next cycle.
+
+The length of the cycle (32 slots) is indicated by the arrows at the bottom of the figure.
+The stations that collide double their contention window, the deterministic backoff after successes and the number of transmitted packet in each transmission attempt.
 

### 2013-03-06 paper.tex
make differences
Azadeh s changes

--- 2013-03-01 paper.tex
@@
 
-\newcommand{\Az}[1]{{\bf{#1}}}
+%\newcommand{\Az}[1]{{\bf{#1}}}
+%\newcommand{\Az}[1]{{\color{blue}{#1}}}
+\newcommand{\Az}[1]{{\color{black}{#1}}}
+%\newcommand{\AzCom}[1]{{\it \color{magenta} [#1]}}
+\newcommand{\AzCom}[1]{}
+%\newcommand{\AzDel}[1]{{\color{red}{#1}}}
+\newcommand{\AzDel}[1]{}
 
@@
 
-A suitable candidate for the replacement of CSMA/CA, should
+A suitable candidate to replace CSMA/CA, should
 \begin{itemize}
@@
 In fact, these proposals are, generally speaking, less aggressive than the currently implemented CSMA/CA protocol.
-Consequently, in a hypothetical mixed network in which the new and old protocols coexisted, the new stations would receive a smaller share of the available bandwidth.
+Consequently, in a hypothetical mixed network in which the new and old protocols coexist, the new stations receive a smaller share of the available bandwidth.
 
@@
 Furthermore, under certain conditions, a collision-free operation can be reached.
-{We will refer hereafter to the class of algorithms that use deterministic backoffs after successful transmissions as CSMA with Enhanced Collision Avoidance (CSMA/ECA).} This approach was first introduced in \cite{barcelo2008lba}, and later a more detailed analysis for both saturated and non-saturated conditions was presented in \cite{bellalta2009vtc}.
+{We will refer hereafter to the class of algorithms that use deterministic backoffs after successful transmissions as CSMA with Enhanced Collision Avoidance (CSMA/ECA).} This approach was first introduced in \cite{barcelo2008lba}, and later a more detailed analysis for both saturated and unsaturated conditions was presented in \cite{bellalta2009vtc}.
 A more in-depth study is carried out in \cite{he2009srb}, including realistic elements such as the possibility of Clear Channel Assessment errors.
@@
 
-Quality-of-service and traffic differentiation in CSMA/ECA has is discussed in \cite{barcelo2009tpc, he2009srb}.
+Quality-of-service and traffic differentiation in CSMA/ECA is discussed in \cite{barcelo2009tpc, he2009srb}.
 The performance of CSMA/ECA in the presence of channel errors is considered in \cite{he2009srb, barcelo2010fcc, fang2011dlm, barcelo2012mdc}.
-Other aspects such as coexistence with legacy stations and performance in the presence of new entrants have also been covered in \cite{he2009srb, barcelo2010fcc, fang2011dlm}.
+Other aspects such as coexistence with legacy stations and the performance when nodes can arbitrarily enter and leave the network have been covered in \cite{he2009srb, barcelo2010fcc, fang2011dlm}.
 Results for the impact of slot drift (non-ideal clocks) in protocols similar to CSMA/ECA can be found in \cite{gong2012asd}.
-In the present workshop paper, we focus on a simple greenfield scenario with a single traffic class, ideal channel, ideal clocks, and saturation conditions.
-We are currently evaluating more complex scenarios that we outline in Section \ref{sec:future_work}
 
@@
 
-The same principles that we exploit to prevent collisions in WLANs can be used in other areas of radio resource management in wireless area networks \cite{duffy2011dcs,checco2012scs,checco2012lbc,khan2013aso}.
+\AzCom{Greenfield paragraph moved to the Results section} The same principles that we exploit to prevent collisions in WLANs can be used in other areas of radio resource management in wireless area networks \cite{duffy2011dcs,checco2012scs,checco2012lbc,khan2013aso}.
 
@@
 The addition of hysteresis and fair-share to the basic CSMA/ECA may represent less frequent transmissions and more packets per transmission
-These two modifications do not interfere in any way with the execution of the legacy protocol and, therefore, CSMA/CA with hysteresis and fair-share can also coexist with the currently deployed protocol.
+These two modifications do not interfere in any way with the execution of the legacy protocol; therefore, CSMA/CA with hysteresis and fair-share can also coexist with the currently deployed protocol.
 
@@
 
-Algorithm \ref{alg:CSMA_CA} describes the CSMA/CA algorithm that is used in current networks.
+Algorithm \ref{alg:CSMA_CA} describes the CSMA/CA protocol that is used in current networks.
 When a station joins the contention, it initializes the retry attempt counter $r$ and the backoff stage $s$ to zero.
@@
 Note that the pseudo-code performs the same action in lines $18$ and $21$.
-Therefore, the algorithm could have been further simplified if instead this action were moved to immediately after line $16$.
+Therefore, the algorithm could be further simplified if instead this action were moved to immediately after line $16$.
 However, the current presentation of the algorithm eases the comparison between the CSMA/CA and the enhancements that follow.
@@
 Firstly, a deterministic backoff is used after successful transmissions.
-This simple modification converts CSMA/CA into CSMA/ECA.
+This simple modification converts CSMA/CA into basic CSMA/ECA.
 Secondly, the backoff stage is not reset after a packet is serviced.
@@
 The only change with respect to CSMA/CA is in fact in line $18$, where the random assignment of $b$ in CSMA/CA is replaced by a deterministic assignment in CSMA/ECA.
-Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2$.
+Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2-1$.
 This value is roughly equal to the expectation of the backoff chosen in line $18$ in Algorithm \ref{alg:CSMA_CA}.
-This particular choice guarantees fairness between new stations and legacy stations.
+This particular choice improves fairness between new stations and legacy stations.
+\AzCom{Does it?
+Even if they are not using aggregation?}
+
+\begin{algorithm}[ht!!!]
+\While{the device is on}
+{
+  %\tcc{Initialize retransmission attempts.}
+  $r \leftarrow 0$; $s \leftarrow 0$\;
+  %\tcc{Initialize backoff counter.}
+  $b \leftarrow \mathcal{U}[0,2^s{\rm{CW}_{min}}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot\;
+        $b \leftarrow b-1$\;
+      }
+      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
+      \If{collision}{
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$\;
+        $s \leftarrow \min (s+1,S)$\;
+        $b \leftarrow \mathcal{U}[0, 2^s {\rm{CW}_{min}} -1]$\;
+      }
+    }
+    $r \leftarrow 0$\;
+    \colorbox{yellow}{$s \leftarrow 0$;}\\
+	 \eIf{success}{
+      %\tcc{Random backoff.}
+      \colorbox{yellow}{$b \leftarrow \mathcal{U}[0,2^{s}{\rm{CW}_{min}}-1]$;}\\
+    }
+    {
+      Discard packet\;
+      $b \leftarrow \mathcal{U}[0,2^s {\rm{CW}_{min}}-1]$\;
+    }
+  }
+  Wait until there is a packet to transmit\;
+}
+\caption{CSMA/CA}
+\label{alg:CSMA_CA}
+\end{algorithm}
+
+\begin{algorithm}[ht!!!]
+\While{the device is on}
+{
+  $r \leftarrow 0$ ; $s \leftarrow 0$\;
+  $b \leftarrow \mathcal{U}[0,2^s{\rm{CW}_{min}}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot\;
+        $b \leftarrow b-1$\;
+      }
+      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
+      \If{collision}
+      {
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$\;
+        $s \leftarrow \min (s+1,S)$\;
+        $b \leftarrow \mathcal{U}[0, 2^s {\rm{CW}_{min}} -1]$\;
+      }
+    }
+    $r \leftarrow 0$\;
+    \colorbox{yellow}{$s \leftarrow 0$;}\\
+    \eIf{success}{
+      %\tcc{Random backoff.}
+      \colorbox{yellow}{$b \leftarrow (2^{s}{\rm{CW}_{min}})/2-1$;}\\
+    }
+    {
+      Discard packet\;
+      $b \leftarrow \mathcal{U}[0, 2^s {\rm{CW}_{min}}-1]$\;
+    }
+  }
+  Wait until there is a packet to transmit\;
+}
+\caption{Basic CSMA/ECA}
+\label{alg:CSMA_ECA}
+\end{algorithm}
+
+\begin{algorithm}[ht!!!]
+\While{the device is on}
+{
+  $r \leftarrow 0$ ; $s \leftarrow 0$\;
+  $b \leftarrow \mathcal{U}[0,2^s{\rm{CW}_{min}}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot\;
+        $b \leftarrow b-1$\;
+      }
+      \colorbox{yellow}{Attempt transmission of 1 packet;}\\
+      \If{collision}{
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$\;
+        $s \leftarrow \min (s+1,S)$\;
+        $b \leftarrow \mathcal{U}[0, 2^s {\rm{CW}_{min}} -1]$\;
+      }
+    }
+    $r \leftarrow 0$\;
+    %$s \leftarrow 0$\;
+    \eIf{success}{
+      %\tcc{Random backoff.}
+      \colorbox{yellow}{$b \leftarrow (2^{s}{\rm{CW}_{min}})/2-1$;}\\
+    }
+    {
+      Discard packet\;
+      $b \leftarrow \mathcal{U}[0, 2^s {\rm{CW}_{min}}-1]$\;
+    }
+  }
+  Wait until there is a packet to transmit\;
+}
+\vspace{0.2cm}
+\caption{CSMA/ECA with hysteresis}
+\label{alg:CSMA_ECA_hist}
+\end{algorithm}
+
+\begin{algorithm}[ht!!!]
+\While{the device is on}
+{
+  $r \leftarrow 0$ ; $s \leftarrow 0$\;
+  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
+  \While{there is a packet to transmit}{
+    %\tcc{Initialize $a$.}
+    \Repeat{($r = R$) or (success)}{
+      %\tcc{First, backoff.}
+      \While{$b>0$}{
+        wait 1 slot\;
+        $b \leftarrow b-1$\;
+      }
+      \colorbox{yellow}{Attempt transmission of $2^s$ packets;}\\
+      \If{collision}{
+        %\tcc{Random backoff.}
+        $r \leftarrow r+1$\;
+        $s \leftarrow \min (s+1,S)$\;
+        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
+      }
+    }
+    $r \leftarrow 0$\;
+    %$s \leftarrow 0$\;
+    \eIf{success}{
+      %\tcc{Random backoff.}
+      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2-1$;}\\
+    }
+    {
+      Discard packet\;
+      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
+    }
+  }
+  Wait until there is a packet to transmit\;
+}
+\vspace{0.2cm}
+\caption{CSMA/ECA with hysteresis and fair-share}
+\label{alg:CSMA_ECA_hist_fair}
+\end{algorithm}
 
@@
 
-\begin{algorithm}[ht!!!]
-\While{the device is on}
-{
-  %\tcc{Initialize retransmission attempts.}
-  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
-  %\tcc{Initialize backoff counter.}
-  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
-  \While{there is a packet to transmit}{
-    %\tcc{Initialize $a$.}
-    \Repeat{($r = R$) or (success)}{
-      %\tcc{First, backoff.}
-      \While{$b>0$}{
-        wait 1 slot \;
-        $b \leftarrow b-1$ \;
-      }
-      \colorbox{yellow}{Attempt transmission of 1 packet } \;
-      \If{collision}{
-        %\tcc{Random backoff.}
-        $r \leftarrow r+1$ \;
-        $s \leftarrow \min (s+1,S)$ \;
-        $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min} -1]$\;
-      }
-    }
-    $r \leftarrow 0$ \;
-    \colorbox{yellow}{$s \leftarrow 0$ } \;
-	 \eIf{success}{
-      %\tcc{Random backoff.}
-      \colorbox{yellow}{$b \leftarrow \mathcal{U}[0,2^{s}\rm{CW}_{min}-1]$} \;
-    }
-    {
-      Discard packet\;
-      $b \leftarrow \mathcal{U}[0,2^s \rm{CW}_{min}-1]$\;
-    }
-  }
-  Wait until there is a packet to transmit \;
-}
-\caption{CSMA/CA}
-\label{alg:CSMA_CA}
-\end{algorithm}
+An example of the operation of CSMA/ECA with hysteresis is presented in Fig. \ref{fig:csma_eca_different_backoff} for a network of $4$ contending stations.
+For each station the backoff counter value at each station is indicated in every slot and the transmissions are represented by rounded rectancles.
+\Az{In the first slot shown, all stations are in the first backoff stage ($s = 0$, $\text{CW}_{\min} = 16$) and have chosen a random backoff value $b \sim \mathcal{U}[0,15]$.
+Stations $3$ and $4$ happen to have the same backoff counter value, and when their counter reaches zero, their transmissions collide.
+Therefore, they increase their backoff stage ($s = 1$) and select another random backoff value accordingly ($b \sim \mathcal{U}[0,31]$).
+On the other hand, stations $1$ and $2$ manage to successfully transmit their packets and move to the deterministic behavior for their next transmission, both selecting a deterministic value equal to $7 = 2^s \rm{CW}_{min}/2 - 1$.
+The first slot in which all stations have different backoff values, marked in the figure with a blue dashed line, can be viewed as the beginning of the collision-free operation.
+Since in this case no two stations will transmit simultaneously, there will be no collision and, by switching to the deterministic behavior, they will all periodically transmit in the same relative position.
+Note that stations $3$ and $4$ move to deterministic behavior when they are in backoff stage $s=1$.
+Therefore, they both use $15 = 2^s \rm{CW}_{min}/2 - 1$ as their deterministic backoff value and send two packets at each transmission opportunity (fair-share).
+The length of the deterministic cycle for the network in this example is $32$ slots, the least common multiple of all stations' cycles.}
+% The stations initially pick a random backoff.}
+% Those stations with the same backoff value collide and choose a new backoff value.
+% When all the stations have a different backoff value, no more collisions will occur.
+% At this point the behaviour of the system is cyclic and deterministic.
+% Note that the state of the system at the beginning of the cycle is the backoff counter of the four stations: 4, 9, 14 and 2.
+% It is exactly the same state that the first slot of the next cycle.
 
-\begin{algorithm}[ht!!!]
-\While{the device is on}
-{
-  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
-  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
-  \While{there is a packet to transmit}{
-    %\tcc{Initialize $a$.}
-    \Repeat{($r = R$) or (success)}{
-      %\tcc{First, backoff.}
-      \While{$b>0$}{
-        wait 1 slot \;
-        $b \leftarrow b-1$ \;
-      }
-      \colorbox{yellow}{Attempt transmission of 1 packet } \;
-      \If{collision}
-      {
-        %\tcc{Random backoff.}
-        $r \leftarrow r+1$ \;
-        $s \leftarrow \min (s+1,S)$ \;
-        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
-      }
-    }
-    $r \leftarrow 0$ \;
-    \colorbox{yellow}{$s \leftarrow 0$} \;    
-    \eIf{success}{
-      %\tcc{Random backoff.}
-      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
-    }
-    {
-      Discard packet\;
-      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
-    }
-  }
-  Wait until there is a packet to transmit \;
-}
-\caption{CSMA/ECA}
-\label{alg:CSMA_ECA}
-\end{algorithm}
+% The length of the cycle (32 slots) is indicated by the arrows at the bottom of the figure.
+% The stations that collide double their contention window, the deterministic backoff after successes and the number of transmitted packet in each transmission attempt.
 
-\begin{algorithm}[ht!!!]
-\While{the device is on}
-{
-  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
-  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
-  \While{there is a packet to transmit}{
-    %\tcc{Initialize $a$.}
-    \Repeat{($r = R$) or (success)}{
-      %\tcc{First, backoff.}
-      \While{$b>0$}{
-        wait 1 slot \;
-        $b \leftarrow b-1$ \;
-      }
-      \colorbox{yellow}{Attempt transmission of 1 packet } \;
-      \If{collision}{
-        %\tcc{Random backoff.}
-        $r \leftarrow r+1$ \;
-        $s \leftarrow \min (s+1,S)$ \;
-        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
-      }
-    }
-    $r \leftarrow 0$ \;
-    %$s \leftarrow 0$ \;
-    \eIf{success}{
-      %\tcc{Random backoff.}
-      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
-    }
-    {
-      Discard packet\;
-      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
-    }
-  }
-  Wait until there is a packet to transmit \;
-}
-\vspace{0.2cm}
-\caption{CSMA/ECA with hysteresis}
-\label{alg:CSMA_ECA_hist}
-\end{algorithm}
-
-\begin{algorithm}[ht!!!]
-\While{the device is on}
-{
-  $r \leftarrow 0$ ; $s \leftarrow 0$ \;
-  $b \leftarrow \mathcal{U}[0,2^s\rm{CW}_{min}-1]$\;
-  \While{there is a packet to transmit}{
-    %\tcc{Initialize $a$.}
-    \Repeat{($r = R$) or (success)}{
-      %\tcc{First, backoff.}
-      \While{$b>0$}{
-        wait 1 slot \;
-        $b \leftarrow b-1$ \;
-      }
-      \colorbox{yellow}{Attempt transmission of $2^s$ packets } \;
-      \If{collision}{
-        %\tcc{Random backoff.}
-        $r \leftarrow r+1$ \;
-        $s \leftarrow \min (s+1,S)$ \;
-        $b \leftarrow \mathcal{U}[0, 2^s  \rm{CW}_{min} -1]$\;
-      }
-    }
-    $r \leftarrow 0$ \;
-    %$s \leftarrow 0$ \;
-    \eIf{success}{
-      %\tcc{Random backoff.}
-      \colorbox{yellow}{$b \leftarrow (2^{s}\rm{CW}_{min})/2$} \;
-    }
-    {
-      Discard packet\;
-      $b \leftarrow \mathcal{U}[0, 2^s \rm{CW}_{min}-1]$\;
-    }
-  }
-  Wait until there is a packet to transmit \;
-}
-\vspace{0.2cm}
-\caption{CSMA/ECA with hysteresis and fair-share}
-\label{alg:CSMA_ECA_hist_fair}
-\end{algorithm}
-
-An example of the operation of CSMA/ECA with hysteresis is presented in Fig. \ref{fig:csma_eca_different_backoff}.
-
-\begin{figure*}[!t]
+\begin{figure*}[!!!htb]
 \centering
 \includegraphics[width=7.0in]{figures/csma_eca_different_backoff}
-\caption{Schedule length distributed adaption example. All the stations transmit the same number of packets in each cycle, despite using different schedule lengths.}
+\caption{An example of contention among CSMA/ECA stations with hysteresis and fair-share.}
 \label{fig:csma_eca_different_backoff}
@@
 
+%\AzDel{In the present workshop paper, we focus on a simple greenfield scenario with
+%a single traffic class, ideal channel, ideal clocks, and saturation conditions.
+%We are currently evaluating more complex scenarios that we outline in Section \ref{sec:future_work}.} \AzCom{Remove?! Can't really fit it in!}
+
 In this section, the performance of the different CSMA protocols is evaluated through simulations.
 The presented results show the saturation throughput and fairness (Jain's Fairness Index (JFI) \cite{jain1984quantitative}) when the number of contenders increases.
-Additionally, the evolution of the probabilities that one slot collision is plotted to provide further insights on the operation of each protocol, and specially whether the collision-free operation is reached.
+Additionally, the evolution of the probability that a slot contains a collision is plotted to provide further insights on the operation of each protocol, and specially whether the collision-free operation is reached.
 
@@
 To evaluate the four protocols, a network of $N$ nodes is considered, where each node is within the coverage area of the others.
-The channel does not introduce errors and the nodes are set to be under saturation (always have packets to transmit).
+The channel does not introduce errors and the nodes are set to be in saturation (always have packets to transmit).
 The number of contenders, $N$, ranges from $2$ to $50$, and one thousand instances of the simulation are performed for each $N$.
@@
 As can be observed, the throughput for CSMA/CA decreases with the number of nodes, as the binary exponential backoff reduces the number of transmission attempts to keep the collisions low.
-The basic CSMA/ECA is able to achieve the collision-free operation if the number of nodes is lower than the deterministic backoff value, i.e., $N\leq \text{CW}_{\min}/2$.
+The basic CSMA/ECA is able to achieve the collision-free operation if the number of nodes is lower than the deterministic cycle length, i.e., $N\leq \text{CW}_{\min}/2$ \AzCom{This time the $-1$ is not required, but it should be cycle length instead of backoff value, right?}.
 This is why in this figure, a phase transition is observed at $N=8$.
@@
 CSMA/ECA with Hysteresis exactly addresses this issue by simply not resetting the backoff stage after a successful transmission.
-In other words, the deterministic backoff value ($2^s \text{CW}_{\min}/2$) is chosen based on the backoff stage, $s$, in which the transmission was successful.
+In other words, the deterministic backoff value ($2^s \text{CW}_{\min}/2-1$) is chosen based on the backoff stage, $s$, in which the transmission was successful.
 This way, a larger number of contenders can reach a collision-free operation (Figure \ref{Fig:throughput}), but at the expense of a lower throughput for $N\leq 8$ compared to the basic CSMA/ECA.
@@
 
-To address the aforementioned fairness issue when hysteresis is used, CSMA/ECA with hysteresis and fair-share was introduced.
-It achieves greater throughput than CSMA/ECA with hysteresis only, while being fair for any number of contenders.
-This is achieved by using packet aggregation, transmitting $2^s$ packets per attempt.
-As observed in Figure \ref{Fig:throughput}, using aggregation we are not only able to provide a fair access, but also we are able to significantly increase the throughput.
-{Note that the throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network.}
+To address the aforementioned fairness issue when hysteresis is used, CSMA/ECA with hysteresis and fair-share was introduced, \Az{whereby stations that access the channel less frequently use packet aggregation to send more (precisely, $2^s$) packets per attempt.} As observed in Figure \ref{Fig:throughput}, using aggregation we are able not only to provide fair access, but also to significantly increase the throughput.
+{The throughput increases with the number of nodes because for larger $N$, the collision-free operation is reached at higher backoff stages, which means more packets will be aggregated and transmitted at every attempt, thus improving the efficiency of the network.}
 
@@
 
-\section{Future Work}
+\section{Conclusions and Future Work}
 \label{sec:future_work}
 
-In this workshop paper, we have evaluated the protocols under discussion in a very simple scenario using ideal channel, ideal clocks and saturation conditions.
+In this paper we have described a suitable replacement of the prevalent CSMA/CA protocol used in the IEEE 802.11 standard.
+Compared to the CSMA/CA protocol, the proposed protocol, called CSMA/ECA with Hysteresis and Fair-Share, offers performance boost in terms of throughput, while preserving long-term fairness.
+In fact, under certain conditions, collision-free operation can be reached in a distributed and adaptive manner, even as the number of nodes in the network increases.
+Furthermore, this protocol is designed with backward compatibility in mind, so that it can operate with legacy nodes without degrading their performance.
+All these properties make the proposed protocol a good candidate to replace CSMA/CA in the upcoming revisions of the standard.
+
+\Az{We have evaluated the protocols under discussion in a very simple scenario using ideal channel, ideal clocks, and in saturation conditions.
 Furthermore, all the participating stations use the same protocol and have a single traffic  class.
-The presented metrics include only throughput and fairness, whereas delay, jitter and packet have been omitted.
-Another scenario of interest that we have not included is the situation in which the stations join the contention to transmit a file and then they leave.
+Our proposed protocol is based on CSMA/ECA, which is shown \cite{?} to work well in non-ideal conditions, and, although not presented in this manuscript, simulation results confirm that CSMA/ECA with hysteresis and fair-share inherits the flexibility and robustness of its predecessors.
 
-Our proposed protocol is based on CSMA/ECA, which works well in non-ideal conditions, and the results we are currently obtaining confirm that CSMA/ECA with hysteresis and fair-share inherits the flexibility and robustness of its predecessors.
+The proposed protocol has yet to be implemented in real prototypes.
+However, current IEEE 802.11 commercial network interfaces do not allow for setting the backoff to a deterministic value after successful transmissions, despite recent efforts on creating firmware that enables higher degree of manipulation  \cite{tinnirello2012wmp}.
+An alternative is using radio-frequency identification (RFID) devices, where there is a programmable hardware that allows for implementing arbitrary protocols.
+RFID uses also a contention protocol which is prone to collisions, and we believe it can be a first arena in which we will be able to develop working prototypes to validate some of the ideas proposed in the present paper.}
 
-Regarding the implementation in real prototypes, it has been proven difficult until now, as the IEEE 802.11 protocols are embedded in the network interfaces and it is not possible to reprogram them.
-There are current advances in that direction \cite{tinnirello2012wmp} but it is still difficult to set the backoff to a deterministic value after successful transmissions in available hardware.
+%\section{Future Work}
 
-Our intention is to try our protocols in a different technology, namely radio-frequency identification (RFID), where there is programmable hardware that can be used to implement arbitrary protocols.
-RFID uses also a contention protocol which is prone to collisions, and we believe it can be a first area in which we will be available to develop working prototypes to validate some of the ideas proposed in the present paper.
-
-\section{Conclusion}
-
-In this paper we have described a suitable replacement of the prevalent CSMA/CA protocol used in the IEEE 802.11 standard.
-Compared to the CSMA/CA protocol, the proposed protocol, called CSMA/ECA with Hysteresis and Fair-Share, offers performance boost in terms of throughput while preserving long-term fairness.
-In fact, under certain conditions, collision-free operation can be reached in a distributed and adaptive manner, even as the number of nodes in the network increases.
-Furthermore, this protocol is backward compatible, and can operate with nodes using the legacy protocol without degrading the performance of the legacy nodes.
-All these properties make the proposed protocol a good candidate to replace CSMA/CA in the upcoming revisions of the standard.
+%\AzCom{Revise from here on!}
+%We have evaluated the protocols under discussion in a very simple scenario using ideal channel, ideal clocks, and in saturation conditions.
+%Furthermore, all the participating stations use the same protocol and have a single traffic  class.
+%The presented metrics include only throughput and fairness, whereas delay, jitter, and packet-loss have not been evaluated.
+%Another scenario of interest not included here is when the stations join the contention only to transmit a file and leave.
+%
+%Our proposed protocol is based on CSMA/ECA, which works well in non-ideal conditions, and the results we are currently obtaining confirm that CSMA/ECA with hysteresis and fair-share inherits the flexibility and robustness of its predecessors.
+%
+%Regarding the implementation in real prototypes, it has been proven difficult until now, as the IEEE 802.11 protocols are embedded in the network interfaces and it is not possible to reprogram them.
+%There are current advances in that direction \cite{tinnirello2012wmp} but it is still difficult to set the backoff to a deterministic value after successful transmissions in available hardware.
+%
+%Our intention is to try our protocols in a different technology, namely radio-frequency identification (RFID), where there is programmable hardware that can be used to implement arbitrary protocols.
+%RFID uses also a contention protocol which is prone to collisions, and we believe it can be a first area in which we will be available to develop working prototypes to validate some of the ideas proposed in the present paper.
 

### 2013-03-06 paper.tex
made differences

@@
 Furthermore, all the participating stations use the same protocol and have a single traffic  class.
-Our proposed protocol is based on CSMA/ECA, which is shown \cite{?} to work well in non-ideal conditions, and, although not presented in this manuscript, simulation results confirm that CSMA/ECA with hysteresis and fair-share inherits the flexibility and robustness of its predecessors.
+Our proposed protocol is based on CSMA/ECA, which is shown \cite{he2009srb,barcelo2010fcc} to work well in non-ideal conditions, and, although not presented in this manuscript, simulation results confirm that CSMA/ECA with hysteresis and fair-share inherits the flexibility and robustness of its predecessors.
 
@@
 The authors would like to thank the anonymous reviewers for their corrections, helpful comments and suggestions for future work.
-The authors are also thankful to Alessandro Checco for helping in the preparation of the related work section.
 
